\documentclass{article}
\usepackage{amsmath, amsthm, amssymb, amsfonts, mathtools, enumitem, stmaryrd,physics, cancel, tikz-cd, graphicx, float, booktabs, mathrsfs}
\usetikzlibrary{arrows}
\usepackage{geometry}
    \geometry{
        a4paper,
        left = 20mm,
        top = 20mm,
        right = 20mm,
        bottom = 30mm
    }
\setlength{\parindent}{0pt}
\setlength{\parskip}{\baselineskip}%

\theoremstyle{definition}
\newtheorem{problem}{Problem}
\newtheorem{solution}{Solution}
\newtheorem*{example}{Example}
\newtheorem*{exercise}{Exercise}
\newtheorem*{definition}{Definition}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem*{theorem*}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem*{proposition*}{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem*{lemma*}{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem*{corollary*}{Corollary}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem*{remark*}{Remark}
\newtheorem{notation}[theorem]{Notation}
\newtheorem*{notation*}{Notation}
\newtheorem*{example*}{Example}
\newtheorem*{convention*}{Convention}


\title{M503 Non-Commutative Algebra}
\author{Taught by Matthias Strauch \\ Notes written by Thanic Nur Samin}
\date{}

\begin{document}
    \maketitle

    \tableofcontents

    Office Hours: Wednesdays 4:30, RH316.

    \section*{Tuesday, 1/13/2026}
    
    \section{Introduction}

    We are interested in non-commutative algebras of finite dimension over a field.

    Reference: Lorenz Algebra Vol. 2, chapters 28 and following.

    General convention:

    \begin{enumerate}[label=\roman*)]
        \item Rings will always assumed to be associative and unital.
        \item Modules are always left modules [unless stated otherwise]. Even when philosophically it's better to consider them as right modules we will turn them into the opposite ring.
    \end{enumerate} 

    \(R =\) commutative unital ring.

    \(A =\) unital \(R\)-algebra, not necessarily commutative.

    Recall: this means that \(A\) is an \(R\)-module, and one has \(\forall x\in R, \forall a,b\in A, x \cdot (ab) = (xa) \cdot b = a \cdot (xb)\).

    \begin{remark}
        \begin{enumerate}[label=\arabic*)]
            \item \(\varphi : R \to A, x \mapsto x \cdot 1_A\) is a ring homomorphism with image in \(Z(A) = \{ a\in A \mid \forall b\in A, ab = ba \}\). 
            
            \[
                (x \cdot 1) \cdot a = x \cdot (1 \cdot a) = x \cdot a = x \cdot (a \cdot a) = a \cdot (x \cdot 1_A)
            \]

            \item Conversely, if \(\psi : R \to B\) [where \(B\) is any unital ring] is an unital ring homomorphism with \(\operatorname{im} (\psi) \subset Z(B)\), then the multiplication \(x \cdot b \coloneqq \psi(x) b\) gives \(B\) the structure of an \(R\)-module and \(B\) becomes an \(R\)-algebra.
            \item The map \(\psi\) in (1) need not be injective. \(\mathbb{Z} = R, \mathbb{Z} / n\mathbb{Z} = A\) is allowed.
            \item Subalgebras of an algebra contain the unit element by convention.
            \item By an \underline{ideal} in \(A\) we always mean a \(2\)-sided ideal. We have left ideals and right ideals defined in the usual way.
            \item A \underline{division algebra} is an \(R\)-algebra \(A\) such that \(\forall a \in A \setminus \{ 0_A \} \exists b \in A\) such that \(ab = ba = 1_A\). It is often also called a skew field. The center of a division algebra is a field.
            
            Modules over a division algebra will also be called vector spaces.

            \item \(A\text{-Mod}\) denotes the category of (left) \(A\)-modules.
        \end{enumerate} 
    \end{remark}

    \begin{proposition}
        Let \(A\) be an \(R\)-algebra and \(M,N \in A\text{-Mod}\). Set \(H = \operatorname{Hom}_A(M,N)\). Then \(H\) is an \(R\)-module with the following action: \(R \times H \to  H, (x,f) \mapsto [m \mapsto x \cdot f(m)]\). This gives \(H\) the structure of an \(R\)-module. 
        
        We need \(R\) to be commutative, because otherwise \(am \mapsto x f(am) = xa f(m)\) which is not necessarily \(ax f(m)\). 
        
        If \(N = M\) then \(H = \operatorname{End}_A(M)\) is in fact an \(R\)-algebra w.r.t.\ the same module structure and composition of maps as multiplication.
    \end{proposition}

    \begin{proof}
        HW1
    \end{proof}

    \begin{remark}
        For \(M \in A\text{-Mod}\) we can regard it also as a module over \(\operatorname{End}_R(M)\).
    \end{remark}

    \begin{notation}
        \begin{enumerate}[label=\arabic*)]
            \item \(A_\ell = A\) considered as an \(A\)-module by left multiplication \(A \times A_\ell \to A_\ell, (a,b) \mapsto ab\).
            \item \(A^o =\) opposite algebra of \(A = A\), but with multiplication defined by \(A \underset{A^o}{\cdot} b \coloneqq b \underset{A}{\cdot} a\). \(A^o\) is still an \(R\)-algebra. We set \(A_r = (A^o)_\ell\). This is just \(A\) but with \(A^o\)-module structure.
            
            \[
                A^o \times A_r \to A_r, (a,b) \mapsto a \underset{A^o}{\cdot} b = ba
            \]
        \end{enumerate} 
    \end{notation}

    \subsection*{The \(R\)-algebra \(M_n(A)\)}

    \(M_n(A)\) denotes the \(R\)-module of \(n \times n\)-matrices with entries in \(A\).

    Multiplication: \((a_{ij})_{1 \leq i,j\leq n} \cdot (b_j)_{i \leq i,j \leq n} \coloneqq \left( \sum_{k = 1}^n a_{ik}b_{kj} \right)_{1 \leq i,j \leq n}\).
    
    \begin{proposition}
        \begin{enumerate}[label=\roman*)]
            \item The map \(\lambda : A \to \operatorname{End}_R(A), \alpha (a) = [b \mapsto ab]\) is injective and induces an isomorphism of \(R\)-algebras \(A \xrightarrow[\lambda]{\cong}\operatorname{End}_{A^o}(A) \subset \operatorname{End}_R(A)\). 
            \item The map \(\rho : A^o \to \operatorname{End}_R(A), \rho (a) = [b \mapsto ba]\) is injective and induces an isomorphism of \(R\)-algebras \(A^o \xrightarrow[\rho]{\cong} \operatorname{End}_A(A) \subset \operatorname{End}_R(A)\). 
        \end{enumerate} 
    \end{proposition}

    We will discuss this later.

    \begin{proposition}

        \begin{enumerate}[label=\roman*)]
            \item The map \(\operatorname{End}_A(A_{\ell}^{\oplus n}) \xrightarrow{\cong} M_n(A^o)\), given by \(f \mapsto (a_{ij})_{i,j}\) where \(f(e_j) = \sum_{i=1}^n a_{ij}e_i\) is an isomorphism of \(R\)-algebras.
            
            \item The map \(\operatorname{End}_{A^o}(A_r^{\oplus n}) \xrightarrow{\cong} M_n(A)\), given by \(f \mapsto (a_{ij})_{i,j}\) where \(f(e_j) = \sum_{i=1}^n a_{ij}e_i\) is an isomorphism of \(R\)-algebras. 
        \end{enumerate}
    \end{proposition}

    \begin{proof}
        \begin{enumerate}[label=\roman*)]
            \item Suppose \(f \mapsto (a_{ij})_{i,j}, g \mapsto (a_{ij})_{i,j}\). Then \((f \circ g)(e_j) = f(g(e_j)) = f \left( \sum_{i=1}^n b_{ij} e_j \right) = \sum_{i=1}^n b_{ij} f(e_j) = \sum_{i=1}^n b_{ij} \sum_{k=1}^n a_{k_i} e_k = \sum_{k=1}^n \left( \sum_{i=1}^n b_{ij} a_{k_i} \right) e_k = \sum_{k=1}^n \left( \sum_{i=1}^n a_{ki} \underset{A^o}{\cdot} b_{ij} \right) e_k\). Note that \(\sum_{i=1}^n a_{ki} \underset{A^o}{\cdot} b_{ij} \)  is simply the entry in position \((k,j)\) of \((a_{ij}) \cdot (b_{ij})\) where they're elements of \(M_n(A^o)\).
            
            Hence, \(f \circ g \mapsto (a_{ij}) \underset{M_n(A^o)}{\cdot} (b_{ij})\).
            
            \item Is identical.
        \end{enumerate} 

    \end{proof}

    Now we generalize. Let \(N \in A \text{-Mod}\) and consider \(N^n\) [which is a shorthand for \(N^{\oplus n}\)] as an \(A \text{-Mod}\) by diagonal multiplication of \(A\).

    Let \(\iota_j : N \to N^n\) be the inclusion of the \(j\)'th summand, \(x \mapsto (0 , \cdots , 0, x , 0, \cdots , 0)\) where \(x\) is in the \(j\)'th position.

    Let \(\pi_i : N^n \to N\) be the projection onto \(i\)'th direct summand, \((x_1 , \cdots , x_n) \mapsto x_i\).  

    \begin{proposition}
        The map \(\operatorname{End}_A(N^n) \xrightarrow{\cong} M_n(\operatorname{End}_A(N))\) where \(f \mapsto (f_{ij})_{1 \leq i,j \leq n}, f_{ij} = \pi_i \circ f \circ \iota_j\) is an isomorphism of \(R\)-algebras. Concretely, \((x_1 , \cdots , x_n) \mapsto \left(\sum_{j=1}^n f_{ij}(x_j) \right)_{i=1,\cdots ,n}\).

        Moreover setting \(C = \operatorname{End}_A(N), N^{\prime} = N^n, C^{\prime} = \operatorname{End}_A(N^{\prime})\) then the inclusion \(\operatorname{End}_C(N) = \operatorname{End}_{\operatorname{End}_A(N)}(N) \xrightarrow{\Delta} \operatorname{End}_C(N^{\prime}) = \operatorname{End}_{\operatorname{End}_A(N)}(N^n) \cong M_n(\operatorname{End}_C(N))\).

        Here \(\Delta : g \mapsto [(x_1 , \cdots , x_n) \mapsto (g(x_1), \cdots , g(x_n))]\).

        This induces an isomorphism (of \(R\)-algebras) \(\operatorname{End}_C(N) \to \operatorname{End}_{C^{\prime}}(N^{\prime}) = \operatorname{End}_{\operatorname{End}_A(N^n)}(N^n)\). 
    \end{proposition}

    Note that \(\operatorname{End}_A(A_\ell) \cong A^o\), so this is sort of a generalization to the previous proposition. 

    \begin{proof}
        The first assertion is an easy exercise.

        Let \(g\in \operatorname{End}_C(N)\) and let \(\widetilde{g}: N^n \to N^n\) be \((x_1 , \cdots , x_n) \mapsto (g(x_1), \cdots , g(x_n))\). We want to show that \(\widetilde{g}\) commutes with elements of \(C^{\prime}\).

        Let \(f\in C^{\prime} = \operatorname{End}_A (N^{\prime})\). 
        
        \(\widetilde{g} (f(x_1, \cdots , x_n)) = \widetilde{g} \left( \left( \sum_{j=1}^n f_{ij}(x_j) \right)\right)_i = \left( g \left( \sum_{j=1}^n f_{ij}(x_j) \right)  \right) _i = \left( \sum_{j} g(f_{ij}(x_j)) \right)_i\)
        
        \(= f \left( (g(x_1), \cdots , g(x_n)) \right)\).
        
        Here \(f_{ij} \in C\).
        
        Surjectivity of \(\operatorname{End}_C(N) \to \operatorname{End}_{C^{\prime}}(N^{\prime})\): given \(h\in \operatorname{End}_{C^{\prime}}(N^{\prime}) [\subset \operatorname{End}_C (N^{\prime}) \cong M_n(\operatorname{End}_C(N))]\), we need to show that it is diagonal. Which means we have to show that \(\forall i \neq j: \pi_i \circ h \circ \iota_j = 0\) and \(\pi_i \circ h \circ \iota_i = \pi_j \circ h \circ \pi_j\).

        \(h \in \operatorname{End}_{C^{\prime}}(N^n) \implies h\) commutes with \(\iota_j \circ \pi_i : N^n \to N^n\). Note that \(\iota_j \circ \pi_i \in C^{\prime} = \operatorname{End}_A(N^n)\).

        Then \(h \circ \iota_j = \iota_j \circ \pi_i \circ h \circ \iota_i\).
        
        Compose with \(\pi_i\) on the left: \(\pi_i \circ h \circ \iota_j = \underbrace{\pi_i \circ \iota_j}_{=0} \circ \pi_i \circ h \circ \iota_i = 0\). 

        Compose with \(\pi_j\) on the left: \(\pi_j \circ h \circ \iota_j = \pi_i \circ h \circ \iota_i\).

        This shows that the matrix of \(h\) is diagonal with constant entries. Let \(\delta = \pi_1 \circ h \circ \iota_1 \in \operatorname{End}(N)\). Then \(h\) corresponds to the diagonal matrix \(\operatorname{diag}(\delta, \dots, \delta)\).

        It remains to show that \(\delta \in \operatorname{End}_C(N)\). Let \(c \in C\) and consider the element \(\bar{c} \in C^{\prime} \cong M_n(C)\) represented by the scalar matrix \(\operatorname{diag}(c, \dots, c)\). Since \(h \in \operatorname{End}_{C^{\prime}}(N^n)\), \(h\) must commute with \(\bar{c}\).
    
        \(\Delta(\delta) \circ \bar{c} = \bar{c} \circ \Delta(\delta) \implies \operatorname{diag}(\delta c, \dots, \delta c) = \operatorname{diag}(c \delta, \dots, c \delta)\).
        
        Thus \(\delta \circ c = c \circ \delta\) for all \(c \in C\), so \(\delta \in \operatorname{End}_C(N)\). Hence \(h = \Delta(\delta)\), proving surjectivity.
        
        Injectivity is clear, as \(\Delta(g)=0\) implies each component \(g=0\).

    \end{proof}

    \section*{Thursday, 1/15/2026}
    
    \begin{corollary}
        \begin{enumerate}[label=\roman*)]
            \item \(\operatorname{End}_A(A^n) \xrightarrow{\cong} M_n(A^o)\).
            \item \(A \xrightarrow{\cong} \operatorname{End}_{\operatorname{End}_A(A^n)}(A^n)\).
            \item \(A^o \cong \operatorname{End}_{M_n(A)}(A^n)\).
            \item \(Z \left( \operatorname{End}_A(A^n) \right) = Z(A^o) \cdot \operatorname{id}_{A^n} = Z(A) \cdot \operatorname{id}_{A^n}\).
            \item \(Z(M_n(A)) = Z(A) \cdot 1_n\) where \(1_n = n \times n\) identity matrix.
        \end{enumerate} 
    \end{corollary}

    \begin{proof}
        \begin{enumerate}[label=\roman*)]
            \item Follows from 1.7 and 1.5.
            \item Take \(N = A_\ell\) in 1.7 and \(C = \operatorname{End}_A(A) \overset{1.5}{=} A^o\).
            
            \(\operatorname{End}_C(N) = \operatorname{End}_{A^o}(A_\ell) = A \xrightarrow[1.7]{\cong} \operatorname{End}_{\operatorname{End}_A(A^n_\ell)}(A^n_\ell)\).  
        \end{enumerate} 

        iii to v are exercises.
    \end{proof}

    \begin{definition}
        [Anti-isomorphism] \(\alpha\) is an \underline{anti-isomorphism} if it is an isomorphism of \(R\)-modules sending \(1_A\) to \(1_A\) and \(\forall a,b \in A \colon \alpha (ab) = \alpha(b) \alpha(a)\).
    \end{definition}

    \begin{proposition}
        \begin{enumerate}[label=\roman*)]
            \item Supppose \(\exists\) anti-isomorphism \(\alpha : A \to A\). Then \(\alpha\) is an isomorphism \(A \to A^o\).
            \item The map \(M_n(A) \to M_n(A^o)^o\) given by \(x \mapsto x^T\) is an isomorphism of algebras.
            \item If \(A\cong A^o\) [as \(R\)-algebras], then \(M_n(A) \cong M_n(A)^o\). In particular, matrix algebras over fields are isomorphic to their opposite. 
        \end{enumerate} 
    \end{proposition}

    \begin{proof}
        HW1.
    \end{proof}

    \begin{example}
        \begin{enumerate}[label=\arabic*)]
            \item \(\mathbb{H} = \mathbb{R} 1 \oplus \mathbb{R} i \oplus \mathbb{R} j \oplus \mathbb{R} k\) the Hamilton Quaternions are isomorphic to \(\mathbb{H}^o\) has an anti-isomorphism given by \(a + bi + cj + dk \mapsto a - bi - cj - dk\). 
            \item Let \(R\) be a field and \(A = \left\{ \begin{bmatrix}
                a & b \\
                0 & d \\
            \end{bmatrix} : a,b,d\in R \right\}\). Then \(A^o \not\cong A\).
        \end{enumerate} 
    \end{example}

    \begin{definition}
        \begin{enumerate}[label=\arabic*)]
            \item A module \(N\) is called \underline{simple} (or \underline{irreducible}) if \(N\neq 0\) and the only submodules of \(N\) are \(0\) and \(N\).
            \item A module \(N\) is called \underline{semisimple} if it is isomorphic to a direct sum of (not necessarily finitely many) simple modules. Formally, we allow the empty direct sum. As a consequence, \(0\) is a semisimple module. 
            \item A submodule \(N \subset M\) is called \underline{minimal} (resp. \underline{maximal}) if it is minimal (resp. maximal) among all non-zero submodules (resp. among all proper submodules). \(0\) doesn't have any minimal or maximal submodule.
        \end{enumerate} 
    \end{definition}

    \begin{example}
        Let \(R = K\) be a field. The simple \(K\)-modules are just the \(1\)-dimensional vector spaces, and they're all isomorphic to \(K\) itself.

        Recall that every vector space has a basis [assuming axiom of choice]. Therefore, every \(K\)-module is semisimple.

        Same is true for division algebras.
    \end{example}

    \setcounter{theorem}{10}
    % proposition 1.10 missing

    \begin{proposition}
        Let \(M\) be a \underline{finitely generated} (f.g.) \(A\)-module, and \(N_0 \subsetneq M\) a proper submodule. Then \(M\) has a maximal submodule \(N\) containing \(N_0\).
    \end{proposition}

    \begin{proof}
        We use Zorn's Lemma.

        Let \(\mathscr{M} = \{ N \subsetneq M : N \text{ submodule, } N \supseteq N_0 \}\).

        \(N_0 \in \mathscr{M} \implies \mathscr{M} \neq \varnothing\).

        Let \(N_1 \subset N_2 \subset \cdots \subset\) an ascending chain in \(\mathscr{M}\). Define \(\widetilde{N} = \bigcup_{i \geq 1} N_i\).

        Let \(m_1 , m_2 , \cdots , m_r\) be generators of \(M\). If \(\widetilde{N} = M\) then there exists \(i \geq 1\) such that \(m_1 , \cdots , m_r \in N_i\). But that would imply \(N_i = M\). This is a contradiction.

        Now we apply Zorn's Lemma. \(\mathscr{M}\) contains a maximal element \(N\). This \(N\) is maximal among all proper submodules. Therefore, \(N \supset N_0\) is maximal.
    \end{proof}

    \begin{proposition}
        Let \(0\neq N \in A\text{-Mod}\). TFAE:

        \begin{enumerate}[label=\roman*)]
            \item \(N\) is simple.
            \item \(\forall x\in N \setminus \{ 0 \} : Ax = N\).
            \item \(\exists \) maximal (left) ideal \(I \subset A : N \underset{A}{\cong} A / I\). 
        \end{enumerate} 
    \end{proposition}

    \begin{proof}
        Consider the map \(A \twoheadrightarrow N\) given by \(a \mapsto ax\). Let \(I = \operatorname{Ann}_A(x) \xrightarrow{N \text{ simple}} I\) is maximal.
    \end{proof}

    \begin{proposition}
        Let \(n > 0\) and \(V\) be an \(n\)-dimensional vector space over a division algebra \(D\), and set \(A = \operatorname{End}_D(V)\). Regard \(V\) as an \(A\)-module. Then,

        \begin{enumerate}[label=\roman*)]
            \item \(V\) is simple as an \(A\)-module.
            \item \(A_\ell \cong V^n\). In particular \(A_\ell\) is a semisimple \(A\)-module.
        \end{enumerate} 
    \end{proposition}

    \begin{proof}
        \begin{enumerate}[label=\roman*)]
            \item We use the fact that if \(v_1 , \cdots , v_n\) is a basis of \(V\) over \(D\), then the map \(\operatorname{End}_D (V) \to \underbrace{V \oplus \cdots \oplus V}_{n} = V^n\) given by \(a \mapsto (a(v_1), \cdots , a(v_n))\) is a bijection of abelian groups.
            
            It is not necessarily a bijection of \(D\)-vector spaces.
            
            Given any \(v_1 \in V \setminus \{ 0 \} \) and any \(w\in V\), we can extend \(\{ v_1 \}\) to be a basis of \(V\) and find \(a\in A\) such that \(a(v_1) = w\). Then \(Av_1 = V \underset{1.12}{\implies} V\) is simple.

            \item Note that the map \(A = \operatorname{End}_D(V) \to V^n\) in (i) is \(A\)-linear, hence an isomorphism of \(A\)-modules.
        \end{enumerate} 
    \end{proof}

    \begin{remark*}
        Take in 1.13 \(V = D^n\). Then \(\operatorname{End}_D(D^n) \underset{1.8}{\cong} M_n(D^o)\).

        Then, 1.13 says \(M_n(D^o) \cong \underbrace{D^n \oplus \cdots \oplus D^n}_{n}\) where the \(D^n\) are columns of the matrices in the LHS. 
    \end{remark*}

    \begin{definition}
        \begin{enumerate}[label=\arabic*)]
            \item An algebra \(A\) is called \underline{semisimple} if \(A_\ell\) is a semisimple \(A\)-module.
            \item \(A\) is called \underline{simple}, if \(A\neq 0\) and does not contain any (\(2\)-sided) ideals other than \(0\) and \(A\).
        \end{enumerate} 
    \end{definition}

    \begin{example}
        \(M_n(D)\) where \(D\) is a division ring is a semisimple algebra by 1.13. This algebra is also simple.

        Note that not every semisimple algebra is simple. Let \(A\) and \(B\) be semisimple. Exercise: \(A \times B\) with componentwise addition and multiplication is semisimple.

        Let \(K_1 , \cdots , K_n\) be fields (or skew fields). Then \(K_1 \times \cdots \times K_n\) is semisimple.
    \end{example}

    Caution: There are simple algebras which are not semisimple (HW1).

    \begin{proposition}
        [Schur's Lemma]

        Let \(M, N \in A\text{-Mod}\). Set \(H = \operatorname{Hom}_A(M,N)\).

        \begin{enumerate}[label=\roman*)]
            \item If \(M\) is simple, any non-zero \(f\in H\) is injective. If \(N\) is simple, any non-zero \(f\in H\) is surjective.
            \item If \(M, N\) are simple, then \(M\cong N\) or \(H = 0\).
            \item If \(M\) is simple, then \(\operatorname{End}_A(M)\) is a division algebra.
        \end{enumerate} 
    \end{proposition}

    \begin{proof}
        Straightforward.
    \end{proof}

    \begin{lemma}
        Suppose \(M = \sum_{i\in I} N_i\) is the sum of a family \((N_i)_{i\in I}\) of \underline{simple} submodules.
        
        Let \(N \subset M\) be any submodule. Then \(\exists J \subset I\) such that,

        \[
            M = N \oplus \bigoplus_{j\in J} N_j
        \]
    \end{lemma}

    \begin{proof}
        Let \(\mathscr{S} = \left\{ J \subset I : \text{s.t. } M + \sum_{j\in J} N_j = N \oplus \bigoplus_{j\in J} N_j  \right\} \). Note that \(\varnothing \in \mathscr{S} \implies \mathscr{S} \neq \varnothing\).

        If \(J_1 \subset J_2 \subset \cdots\) is a chain in \(\mathscr{S}\) and \(\widetilde{J} = \bigcup_{k=1}^{\infty} J_k\) then [exercise] \(\widetilde{J} \in \mathscr{S}\).

        Zorn's lemma implies that \(\mathscr{S}\) has a maximal element \(J\).

        Check: \(J\) works.

    \end{proof}

    \section*{Tuesday, 1/20/2026}
    
    \begin{proposition}
        For \(M \in A \text{-Mod}\). TFAE:


        \begin{enumerate}[label=\roman*)]
            \item \(M\) is the sum of simple submodules.
            \item \(M\) is semisimple.
            \item Every submodule is a direct summand of \(M\). 
        \end{enumerate} 
    \end{proposition}

    \begin{proof}
        i \(\implies\) ii: take \(N = 0\) in 1.15.

        ii \(\implies\) iii: Apply 1.15 with the submodule as \(N\).

        iii \(\implies\) i: Let \(x\in M \setminus \{ 0 \} \) and consider the `cyclic submodule' \(C = Ax \subset M\). Note that \(x\in C \implies 0\neq C\).

        1.11 implies \(\exists\) maximal submodule \(L \subsetneq C\). Since \(L\) is a maximal submodule, \(C / L\) must be simple.

        By assumption, \(\exists N \subset M\) such that \(M = L \oplus N\). Since \(L \subset C\) it follows that \(C = L \oplus (C \cap N)\).

        Therefore, \(C / L \cong C \cap N \implies C \cap N\) is a simple module.

        It follows that every non-zero submodule of \(M\) contains a simple submodule.

        Set \(M^{\prime} = \sum_{N \subset M \text{ simple}} N\). Let \(M^{\prime\prime} \subset M\) be the summand, i.e. \(M = M^{\prime} \oplus M^{\prime\prime}\).
        
        If \(M^{\prime\prime} \neq 0\), it contains a simple submodule \(N_0 \subseteq M^{\prime\prime}\). But then \(N_0 \subset M^{\prime}\). This is a contradiction.
        
        Thus, \(M^{\prime\prime} = 0\). Which means \(M = M^{\prime} =\) sum of simple submodules.
    \end{proof}

    \begin{remark*}
        Given any \(M \in A\text{-Mod}\), the \underline{socle} of \(M\), \(\operatorname{soc}(M)\) is defined to be the largest semisimple submodule of \(M\).
        
        By 1.16, \(\operatorname{soc}(M) = \sum_{N \subset M \text{ simple}} N\). 
    \end{remark*}

    \begin{proposition}
        Suppose \(M\) is the sum of simple submodules \((N_i)_{i\in I}\). Then,

        \begin{enumerate}[label=\roman*)]
            \item Every submodule or quotient module is isomorphic to \(\bigoplus_{j\in J} N_j\) for some \(J \subset I\). Note that it need not be directly equal, just isomorphic. Consider \(K \xhookrightarrow{\Delta} K \oplus K\).
            \item Any simple submodule \(N \subset M\) is isomorphic to one of the \(N_i\). 
        \end{enumerate} 
    \end{proposition}

    \begin{proof}
        ii follows from i.

        For i: Let \(M \underset{\pi}{\twoheadrightarrow} M^{\prime\prime}\) be a quotient of \(M\). Then, \(\overline{N_i} \coloneqq \pi (N_i)\) is zero or simple.
        
        Set \(\overline{I} \coloneqq \{ i\in I \mid \overline{N_i} \neq 0 \} \).
        
        \(\implies M^{\prime\prime} = \sum_{i\in \overline{I}} \overline{N_i}\). Applying 1.15, \(\exists \overline{J} \subset \overline{I}\) so that \(M^{\prime\prime} = \bigoplus_{j\in \overline{J}} \overline{N_j} \cong \bigoplus_{j\in \overline{J}} N_j\).

        If \(M^{\prime} \subset M\) is a submodule, 1.16 \(\implies \exists N \subset M : M = N \oplus M^{\prime}\). Then \(M^{\prime} \cong M / N\), which is a quotient of \(M\).
    \end{proof}

    \begin{corollary}
        \begin{enumerate}[label=\roman*)]
            \item Every submodule or a quotient module of a semisimple module is semisimple.
            \item If \(A\) is semisimple, any simple \(A\)-module is isomorphic to a submodule of \(A_\ell\), and so is isomorphic to a minimal left ideal of \(A\). 
        \end{enumerate} 
    \end{corollary}

    \begin{proof}
        \begin{enumerate}[label=\roman*)]
            \item Directly follows from 1.17.
            \item  Suppose \(A\) is semisimple. Then by definition, \(A_\ell = \bigoplus_{i\in I} N_i\) where \(N_i\) are simple. If \(N\) is a simple module, 1.12 implies \(N \cong A_\ell / I\) where \(I \subset A\) is a maximal submodule.
            
            1.17 implies that \(N\cong N_i\) for some \(i\).

            Note that each \(N_i\) is a left ideal. Since it is simple, it must be minimal among the non-zero submodules.
        \end{enumerate} 
    \end{proof}

    \begin{proposition}
        For an algebra \(A\) TFAE:

        \begin{enumerate}[label=\roman*)]
            \item \(A\) is semisimple.
            \item Every \(A\)-module is semisimple. 
        \end{enumerate} 
    \end{proposition}

    \begin{proof}
        ii \(\implies\) i is direct.

        i \(\implies\) ii: Suppose \((m_i)_{i\in I}\) is a set of generators of \(M\) as an \(A\)-module. Consider \(A_\ell^{(I)} \coloneqq \bigoplus_{i\in I} A_\ell \twoheadrightarrow M\) given by \((a_i)_{i\in I} \mapsto \sum_{i} a_i m_i\).

        Thus \(M\) must be a quotient of a semisimple module. The statement follows from 1.18.
    \end{proof}

    \begin{definition}
        Denote by \(\mathcal{T}(A)\) the set of isomorphism classes of simple \(A\)-modules.

        We call a simple module \(N\) of type \(\tau \in \mathcal{T}(A)\) if the isomorphism class of \(N\) is \(\tau\).

        Given \(M\in A\text{-Mod}, \tau \in \mathcal{T}(A)\), we set \(M_{\tau} = \sum_{N \subset M \text{ simple of type } \tau} N\) [note that it must be a direct sum of some submodules of type \(\tau\).] \(M_{\tau} \) is called the \underline{\(\tau\)-isotypic component}.

        If \(M = M_{\tau}\) then we say that \(M\) is \underline{isotypic of type \(\tau\).} 
    \end{definition}

    \begin{remark*}
        By 1.12, \(\left\{ I \subset A_\ell \mid \text{maximal left ideal} \right\} \to \mathcal{T}(A)\) given by \(I \mapsto \text{isomorphism class of } A / I\) is surjective. Thus \(\mathcal{T}(A)\) is a set.
        
        Exercise: This map is bijective (HW2).
    \end{remark*}

    \begin{notation*}
        \(A\text{-Mod}^{\text{ss}} =\) category of semisimple \(A\)-modules.
    \end{notation*}

    \begin{corollary}
        Let \(M \in A\text{-Mod}^{\text{ss}}\). Then,
        
        \begin{enumerate}[label=\roman*)]
            \item \(M = \bigoplus_{\tau \in \mathcal{T}(A)} M_{\tau}\).
            \item \(\forall\) submodule \(M^{\prime} \subset M: M^{\prime} = \bigoplus_{\tau \in \mathcal{T}(A)} (M^{\prime} \cap M_{\tau})\).   
        \end{enumerate} 
    \end{corollary}

    \begin{proof}
        \begin{enumerate}[label=\roman*)]
            \item \(M\) semisimple \(\implies M = \sum_{\tau} M_{\tau}\). Fix \(\tau\). 1.18 implies that \(M^{\prime} \coloneqq M_{\tau} \cap \left( \sum_{\tau^{\prime} \neq \tau} M_{\tau^{\prime}} \right)\) is semisimple.
            
            If \(M^{\prime} \neq 0\) then \(\exists N \subset M^{\prime}\) which is simple. \(N \subset M_\tau\) implies \(N\) must be of type \(\tau\). However, \(N \subset \sum_{\tau^{\prime} \neq \tau} M_{\tau^{\prime}}\),, which implies that \(N\) must not be of type \(\tau\). This is a contradiction. Therefore, \(M^{\prime} = 0\).

            This means the sum is direct, i.e. \(M^{\prime} = \bigoplus_{\tau} M_{\tau}\).

            \item 1.18 implies that \(M^{\prime}\) must be semisimple. Therefore \(M^{\prime} = \bigoplus_{\tau} M^{\prime}_{\tau}\).
            
            Clearly \(M_{\tau}^\prime \subset M^{\prime} \cap M_{\tau}\).

            Conversely, \(M^{\prime} \cap M_{\tau}\) is semisimple by 1.18. Therefore, \(M^{\prime} \cap M_{\tau} = \sum_{N \subset M^{\prime} \cap M_{\tau}, N \text{ simple} } N\). Note that each \(N\) is of type \(\tau\). Thus \(M^{\prime} \cap M_\tau = M^\prime_{\tau}\).
        \end{enumerate} 
    \end{proof}

    \begin{proposition}
        If \(M, M^{\prime} \in A \text{-Mod}^{\text{ss}}\), then \(\operatorname{Hom}_A (M,M^{\prime}) = \coprod_{\tau \in \mathcal{T}(A)} \operatorname{Hom}_A(M_{\tau}, M_{\tau}')\). 
    \end{proposition}

    \begin{proof}
        Suppose \(f\in \operatorname{Hom}_A(M, M^{\prime})\). Consider \(\eval{f}_{M_{\tau}}: M_{\tau} \twoheadrightarrow f(M_{\tau}) \subset M^{\prime}\).
        
        Note that \(f(M_{\tau}) = \sum_{N \subset M \text{ simple of type } \tau} f(N)\). Thus \(f(M_{\tau}) \subset M^{\prime}_{\tau}\).  
    \end{proof}

    \begin{proposition}
        \(M \in A \text{-Mod}^{ss}\). For a submodule \(U \subset M\) TFAE:

        \begin{enumerate}[label=\roman*)]
            \item \(\forall f\in \operatorname{End}_A(M), f(U) \subset U\).
            \item \(U\) is a direct sum of some \(M_{\tau}, \tau \in \mathcal{T}(A)\).  
        \end{enumerate}
        
        \begin{proof}
            ii \(\implies\) i is a direct consequence of 1.21.

            i \(\implies\) ii: 1.20 and 1.18 imply \(U = \bigoplus_{\tau} U_{\tau}\).

            If \(U_{\tau} \subsetneq M_{\tau}\), 1.15 \(\implies M_{\tau} = U_{\tau} \bigoplus_{i\in I} N_i\) where the sum is non-zero and each \(N_i\) is of type \(\tau\).

            Then one finds \(f: U \to M\) such that \(f(U) \not\subset U\). Can extend \(f\) to a homomorphism \(f: M \to M\). 
        \end{proof}
    \end{proposition}

    \section*{Thursday, 1/22/2026}
    
    \begin{corollary}
        Let \(A\) be semisimple. For a subset \(U \subseteq A\) TFAE:

        \begin{enumerate}[label=\roman*)]
            \item \(U\) is an ideal of \(A\). (By convention we mean a two-sided ideal). 
            \item \(U\) is a direct sum of isotypic components of \(A_\ell\).
        \end{enumerate} 
    \end{corollary}

    \begin{proof}
        i \(\implies\) ii: \(U\) is an ideal \(\implies U\) is a submodule of \(A_\ell\), and \(U\) is stable under `right multiplication by \(a\)'. We denote this right multiplication by \(a\) map to be \(_A a : A_\ell \to A_\ell\), \(_A a (b) = ba\). So \(U\) is stable under all \(_A a\).
        
        1.5 \(\implies \operatorname{End}_A(A_\ell) = A^o\).
        
        Therefore, \(\forall f\in \operatorname{End}_A(A_\ell) : f(U) \subset U\). Statement ii follows from 1.22.

        ii \(\impliedby\) i: same argument in reverse order.
    \end{proof}

    \begin{corollary}
        Let \(A\) be semisimple.

        \begin{enumerate}[label=\roman*)]
            \item The isotypic components of \(A_\ell\) are precisely the minimal ideals of \(A\), and every ideal is a direct sum of minimal ideals.
            \item \(A\) has only finitely many minimal ideals. 
            \item \(\vert \mathcal{T}(A) \vert < \infty\).
        \end{enumerate} 
    \end{corollary}

    \begin{proof}
        \begin{enumerate}[label=\roman*)]
            \item \(I \subseteq A_\ell\) a direct sum of isotypic components \(\underset{1.23}{\iff} I\) is an ideal.
            
            Therefore, \(I\) is an isotypic component \(\iff\) \(I\) is a minimal ideal.

            \item Write \(A = \bigoplus_{\tau \in \mathcal{T}(A)} (A_\ell)_{\tau}\). Then we can write \(1 = \sum_{\tau} a_{\tau} \) where \(a_{\tau} \in (A_{\ell})_{\tau}\) and all but finitely many \(a_\tau\) are zero.
            
            Then, \(A = \sum_{\tau, a_{\tau} \neq 0} A a_{\tau} \subseteq \bigoplus_{a_{\tau} \neq 0} (A_\ell)_{\tau} \subseteq A\).

            \item 1.18: any simple \(A\)-module \(N\) is isomorphic to a submodule of \(A_\ell\). Suppose \(\tau\) is the type of \(N\). Then, \((A_\ell)_{\tau} \supseteq N\).
            
            Since there are only finitely many isotypic components, there are only finitely many \(\tau \in \mathcal{T}(A)\).
        \end{enumerate} 
    \end{proof}

    \begin{example*}
        \begin{enumerate}[label=\arabic*)]
            \item Let \(K_1 , \cdots , K_n\) be fields. Then, \(A = K_1 \times \cdots \times K_n\) is semisimple, and \(K_i = \{ 0 \} \times \cdots \times \{ 0 \} \times K_i \times \{ 0 \} \times \cdots \times \{ 0 \} \) is a simple module, and these exhaust all of \(\mathcal{T}(A)\).  
            
            \item Suppose \(K\) is a field, and let \(A = M_n(K)\). We can write \(A_\ell\) as a direct sum of columns, i.e. \(A_\ell = Ae_1 \oplus \cdots \oplus Ae_n\).
            
            Here, \(A_{e_i} \underset{A}{\cong} K^n\) is up to isomorphism the only simple \(A\)-module. Denote by \(\tau\) this type. \((A_\ell)_{\tau}\) is the only isotypic component.

            \item \(A = M_{n_1} (K_1) \times \cdots \times M_{n_s}(K_s)\) has \(s\) isotypic components.
        \end{enumerate} 
    \end{example*}

    \begin{lemma}
        Let \(M \in A\text{-Mod}^{\text{ss}}\) and suppose \(M = \bigoplus_{i\in I} N_i\) and \(M = \bigoplus_{j\in J} N_j^{\prime}\) with simple submodules \(N_i , N_j^{\prime} \subset M\).
        
        Then \(\exists\) bijection \(\sigma : I \to J\) such that \(\forall i\in I : N^{\prime}_{\sigma(i)} \underset{A}{\cong} N_i\). In particular, \(I\) and \(J\) must have the same cardinality.
    \end{lemma}

    \begin{proof}
        WLOG we may assume \(M = M_{\tau}\). Let \(N\) be a fixed simple module of type \(\tau\).

        Then \(N \cong N_i \cong N_j^{\prime}\) for all \(i\in I, j \in J\).

        Put \(D = \operatorname{End}_A(N)^o\). 1.44 \(\implies D\) is a division algebra.

        Then, the map \(D \times \operatorname{Hom}_A(N,M) \to \operatorname{Hom}_A(N,M)\) given by \((d,f) \mapsto f \circ d\) makes \(\operatorname{Hom}_A(N,M)\) into a \(D\)-module.
        
        Moreover, \(\operatorname{Hom}_A(N, M) = \bigoplus_{i\in I} \operatorname{Hom}_A (N, N_i)\).
        
        However, \(\operatorname{Hom}_A(N, N_i) \cong \operatorname{Hom}_A(N,N) = \operatorname{End}_A(N) \cong D^o\). Therefore, \(\operatorname{Hom}_A(N, N_i)\) is free of rank \(1\) over \(D\).
        
        Then, \(\dim_D \operatorname{Hom}_A(N, M) = \vert I \vert\).

        By the same argument, \(\dim_D \operatorname{Hom}_A(N, M) = \vert J \vert\) using the other decomposition.
    \end{proof}

    \begin{definition}
        If we write \(M\in A\text{-Mod}^{\text{ss}}\) as \(M = \bigoplus_{i \in I} N_i\) where \(N_i\) are simple, then \(\ell_A(M) = \vert I \vert \) is called the \underline{length} of \(M\) (as an \(A\)-module).
        
        The \underline{length of a semisimple algebra \(A\)} is by definition \(\ell_A(A_\ell)\).

        Given a simple module \(N\) of type \(\tau\) we also use the notation:

        \[
            M : \tau = M : N \coloneqq \ell_A(M_{\tau})
        \]
    \end{definition}

    \begin{remark*}
        If \(N\) is simple and \(D = \operatorname{End}_A(N)^o\), then \(M : N = \dim_D \operatorname{Hom}_A(N,M)\) and \(\ell_A(M) = \sum_{\tau \in \mathcal{T} (A)} M : \tau\) 
    \end{remark*}

    \begin{theorem}
        [Jacobson Density Theorem]

        Let \(M\in A\text{-Mod}^{\text{ss}}\) and \(C = \operatorname{End}_A(M)\) and consider the canonical map:

        \begin{enumerate}[label=(\arabic*)]
            \item \(A \to \operatorname{End}_C(M), a \mapsto a_M = [m \mapsto am]\). Note that this map is not necessarily \(A\)-linear. But it is \(C\)-linear.
        \end{enumerate} 

        Then, the image of (1) is `dense' in \(\operatorname{End}_C(M)\) in the following sense:

        \(\forall f\in \operatorname{End}_C(M)\) and \(\forall x_1 , \cdots , x_n \in M, \exists a \in A\) such that \(\forall 1 \leq i \leq n : f(x_i) = a \cdot x_i \quad (2)\).

        If \(M\) is finitely generated as a \(C\)-module, the map (1) is surjective.
    \end{theorem}

    \begin{proof}
        Consider \(M^{\prime} \coloneqq M^{\oplus n} \in A\text{-Mod}^{\text{ss}}\) and \(x \coloneqq (x_1 , \cdots , x_n) \in M^{\prime}\).
        
        1.16 \(\implies \exists M^{\prime\prime} \subset M^{\prime}\) with the property \(M^{\prime} = M^{\prime\prime} \oplus Ax\).

        Then we can define a projection \(p\) as follows: \(p: M^{\prime} \to M^{\prime}\) as follows: every element of \(M^{\prime} = M^{\prime\prime} + Ax\) can be written as \(m^{\prime\prime} + ax\) where \(m^{\prime\prime} \in M^{\prime\prime}\). Then, \(p(m^{\prime\prime} + ax) \coloneqq ax \in M^{\prime}\).

        Then, \(p\in \operatorname{End}_A(M^{\prime}) \eqqcolon C^{\prime}\).
        
        Now define \(\widetilde{f}: M^{\prime} \to M^{\prime}\) as follows: \(\widetilde{f}(y_1 , \cdots , y_n) = (f(y_1) , \cdots , f(y_n))\).

        1.7 \(\implies \operatorname{End}_C (M) \xrightarrow{\cong} \operatorname{End}_{C^{\prime}}(M^{\prime})\) given by \(f \mapsto \widetilde{f}\) is a \(C^{\prime}\)-linear bijection.

        Thus, \(\widetilde{f} \circ p = p \circ \widetilde{f}\).

        Therefore, \((f(x_1), \cdots , f(x_n)) = \widetilde{f}(x) = \widetilde{f}(p(x)) = p(\widetilde{f}(x)) = a \cdot x\) for some \(a\in A\) [since \(\operatorname{im} p = Ax\)].

        Therefore, \(\forall 1 \leq i \leq n : f(x_i) = a x_i\).
    \end{proof}

    \begin{corollary}
        Let \(N\in A\text{-Mod}\) be simple and \(D = \operatorname{End}_A(N)\). Consider \(N\) as a \(D\)-vector space. Let \(x_1 , \cdots , x_n\) be linearly independent over \(D\). Then for any \(y_1 , \cdots , y_n \in N \exists a \in A \forall 1 \leq i \leq n : y_i = a x_i\). 
    \end{corollary}

    \begin{proof}
        We apply 1.26 with \(C = \operatorname{End}_A(N) = D\).

        We extend \((x_1 , \cdots , x_n)\) to a \(D\)-basis of \(N\). Then we can choose \(f \in \operatorname{End}_{D = C} (N)\) such that \(f(x_i) = y_i\) for all \(1 \leq i \leq n\).
    \end{proof}

    \section*{Tuesday, 1/27/2026}
    
    \subsection*{Noetherian and Artinian Modules (E. Artin, 1898-1962, IU 1938-1946)}

    \begin{definition}
        A module \(M\) is called:

        \begin{enumerate}[label=\roman*)]
            \item \underline{Noetherian} if every non-empty subset of submodules of \(M\) has a maximal element.
            \item \underline{Artinian} if every non-empty subset of submodules of \(M\) has a minimal element. 
        \end{enumerate} 

        An algebra \(A\) is called \underline{artinian} (resp. \underline{artinian}) if \(A_\ell\) is noetherian (resp. artinian).
    \end{definition}

    \begin{remark}
        \begin{enumerate}[label=\roman*)]
            \item \(M\) is noetherian (resp. artinian) iff every increasing (resp. decreasing) (countable) chain is stationary (i.e. all members are the same from some index on).
            \item For a division ring \(D\) and \(V\) a \(D\)-Mod, TFAE:
            
            \(V\) is noetherian \(\iff V\) is artinian \(\iff \dim_D(V) < \infty\).

            \item If \(M = \bigoplus_{i\in I} M_i\) with all \(M_i\) non-zero and \(\vert I \vert = \infty\), then \(M\) is neither noetherian nnor artinian.
            \item \(\mathbb{Z}\) is not artinia: \((2) \supsetneq (6) \supsetneq (24) \supsetneq (120) \supsetneq \cdots\). \(\mathbb{Z}\) is noetherian, as is any PID.
            \item \(\mathbb{Q} / \mathbb{Z}\) (as a \(\mathbb{Z}\)-module) is not noetherian: \(\frac{1}{2}\mathbb{Z} / \mathbb{Z} \subsetneq \frac{1}{6}\mathbb{Z} / \mathbb{Z}  \subsetneq \frac{1}{24} \mathbb{Z} / \mathbb{Z} \subsetneq \cdots\).
            
            \(\mathbb{Q} / \mathbb{Z}\) is also not artinian: \(M = \sum_{p \text{ prime}} \frac{1}{p}\mathbb{Z} / \mathbb{Z} \supsetneq \sum_{p > 2} \frac{1}{p} \mathbb{Z}  / \mathbb{Z} \supsetneq \sum_{p > 3} \frac{1}{p}\mathbb{Z} / \mathbb{Z} \supsetneq \cdots\).

            But for \(\mathbb{Z}_{(p)}\) [the localization of \(\mathbb{Z}\) at \(p\)], the module \(\mathbb{Q} / \mathbb{Z}_{(p)}\) is artinian.

            \item If \(K\) is a field, then any finite dimensional \(K\)-algebra \(A\) is noetherian and artinian.
            \item \(\exists\) noetoherian and artiniain \(K\)-algebra \(A\) such that \(A^o\) is neither noetherian nor artinian.
            \item We'll prove: \(A\) artinian \(\implies A\) is noetherian. 
        \end{enumerate} 
    \end{remark}

    \begin{proposition}
        \(M\) is noetherian iff every submodule is finitely generated.
    \end{proposition}

    \begin{proof}
        \(\implies\): given \(N \subset M\) set \(S = \{ N^{\prime} \subset N \mid N^{\prime} \text{ is f.g.} \} \).

        \(0 \in S \implies S \neq \varnothing\), and by assumption \(S\) has a maximal element \(N_0 \subseteq N\). It is easy t show that \(N_0 = N \implies N\) is f.g.

        \(\impliedby\): Let \(N_1 \subseteq N_2 \subseteq \cdots\) be submodules of \(M\). By assumption we can show that \(N = \bigcup_{i} N_i\) is finiely generated. Let \(m_1 , \cdots , m_s\) be generators of \(N\). Then \(\exists i_0: m_1 , \cdots , m_s \in N_{i_0}\). Thus \(N = N_{i_0} = N_{i_0 + 1} = \cdots\). 
    \end{proof}

    \begin{remark*}
        For \(M\) to be noetherian it does not suffice that \(M\) is finitely generated.

        If \(A\) is noetherian and \(M\) is finitely generated, then \(M\) is noetherian.
    \end{remark*}

    \begin{proposition}
        Let \(M\) be a module and \(N \subset M\) be a submodule.

        \begin{enumerate}[label=\roman*)]
            \item \(N\) and \(M / N\) are noetherian \(\iff M\) is noetherian.
            \item \(N\) and \(M / N\) are artinian \(\iff\) \(M\) is artinian. 
        \end{enumerate} 
    \end{proposition}

    \begin{proof}
        HW3
    \end{proof}

    \begin{corollary}
        Let \(p\) be the property `noetherian' or `artinian'. 

        \begin{enumerate}[label=\roman*)]
            \item Suppose \(M = \bigoplus_{i=1}^n M_i\). Then \(M\) is \(p\) iff all \(M_i\) are \(p\).
            \item If an algebra \(A\) is \(p\), every \(M \in A\text{-Mod}^{\text{f.g.}}\) is \(p\).
        \end{enumerate} 
    \end{corollary}

    Suppose,

    \[
        0 \to N \hookrightarrow M \to M / N \to 0
    \]

    say: \(M\) is an extension of \(M / N\) by \(N\).

    \begin{proof}
        \begin{enumerate}[label=\roman*)]
            \item Reduce to the case \(n = 2\) and apply 1.30.
            \item \(A\) is \(p \implies A_\ell\) is \(p \xrightarrow{\text{(i)}} A_\ell^{\oplus n}\) is \(p. 1.30 \implies A_\ell^{\oplus n} / L\) is \(p\) for any submodule \(L \subset A_\ell^{\oplus n}\). Any finitely generated module is of this form which implies the assertion.  
        \end{enumerate} 
    \end{proof}

    \begin{proposition}
        Let \(M \in A\text{-Mod}^{\text{ss}}\). TFAE:

        \begin{enumerate}[label=\roman*)]
            \item \(M\) is finitely generated.
            \item \(M\) is a finite direct sum of simple submodules.
            \item \(M\) is artinian.
            \item \(M\) is noetherian. 
        \end{enumerate} 
    \end{proposition}

    \begin{proof}
        Write \(M = \bigoplus_{i\in I} N_i\) where \(N_i\) are simple. Assume i, and let \(m_1 , \cdots , m_s\) be generators. \(\forall j : 1 \leq j \leq s \exists I_j \subset I\) such that \(\vert I_j \vert < \infty\) such that \(m_j \in \bigoplus_{i\in I_j} N_i\).
        
        Then, \(m_1 , \cdots , m_s \in \bigoplus_{i\in \bigcup_{j=1}^s I_j} N_i = M\) which implies ii.
        
        ii \(\implies\) iii, iv is implied by 1.31.

        iv \(\implies\) i: Noetherian implies finitely generated.

        To complete the full circle, assume iii. Use remark 1.28. HW3.
    \end{proof}

    \begin{corollary}
        Every semisimple algebra \(A\) is noetherian and artinian.
    \end{corollary}

    \begin{proof}
        \(A_\ell\) is semisimple and f.g., \(A_\ell = A \cdot 1_A\). Apply 1.32.
    \end{proof}

    \section*{Thursday, 1/29/2026}
    
    \begin{remark*}
        The study of Semisimple Algebras has very little overlap with Commutative Algebras, namely finite product of fields.

        There is a generalization: \underline{Azumaya algebras}.

        Goro Azumaya: IU (1968-1992)
    \end{remark*}

    \begin{proposition}
        [Definition of Composition Series]
        For \(M\in A\text{-Mod}\) TFAE:

        \begin{enumerate}[label=\roman*)]
            \item \(M\) is both artinian and noetherian.
            \item \(M\) has a \underline{composition series}: a chain of submodules:
            
            \[
                M = M_n \supsetneq M_{n-1} \supsetneq \cdots \supsetneq M_0 \coloneqq 0
            \]

            where \(M_i / M_{i-1}\) is simple for \(1 \leq i \leq n [n=0 \text{ if } M = 0]\).

            \(n\) is called the \underline{length of the composition series}.
        \end{enumerate} 
    \end{proposition}

    \begin{proof}
        ii \(\implies\) i: Suppose \(M\) has a composition series of some length. Then \(M_1\) must be simple. If there is an \(M_2\), \(M_2 / M_1\) must be simple. We have the exact sequence:

        \[
            0 \to M_1 \to M_2 \to M_2 / M_1
        \]

        1.30 then implies that \(M_2\) is noetherian and artinian. The fact that \(M\) must be both noetherian and artinian follows from induction.

        i \(\implies\) ii: HW3.
    \end{proof}

    \begin{theorem}
        [Jordan-H\"older Theorem for Modules]

        If \(M\) has to composition series:

        \begin{enumerate}[label=\arabic*)]
            \item \(M = M_n \supsetneq M_{n-1} \supsetneq \cdots \supsetneq M_0 = 0\).
            \item \(M = L_m \supsetneq L_{m-1} \supsetneq \cdots \supsetneq L_0 = 0\).
        \end{enumerate} 

        then \(m = n\). Furthermore, the simple quotient are the same under permutation: \(\exists \sigma \in \mathcal{S}_n\) such that \(\forall 1 \leq i \leq n: L_i / L_{i-1} \underset{A}{\cong} M_{\sigma(i)} / M_{\sigma(i) - 1}\).
    \end{theorem}

    \begin{definition}
        A module \(M\) is called of \underline{finite length} if \(M\) satisfies the equivalent conditions of 1.34.

        The length of any composition series is called \underline{the length of} \(M\), denoted by \(\ell(M)\).
    \end{definition}

    \begin{remark*}
        \begin{enumerate}[label=\roman*)]
            \item If \(M\) is of finite length and \(N\) is a submodule, then \(N\) is of finite length. Furthermore,
            
            \[
                \ell(M) = \ell(N) + \ell(M / N).
            \]

            We use the exact sequence:

            \[
                0 \to N \to M \to M / N.
            \]

            \item \(M\) semisimple of finite length, i.e. \(M = \bigoplus_{i=1}^n N_i\) where \(N_i\) are simple \(\implies \ell(M) = n\).
            
            \(M\) is not always the direct sum of the simple quotients. For example, suppose \(K\) is a field and \(A = \left\{ n \times n \text{ upper triangular \(K\)-matrices} \right\} \) acting on \(K^n\). Then \(Ke_1\) is a submodule but \(Ke_2\) is not one.
            
            \[
                K^n \supsetneq Ke_1 \oplus \cdots Ke_{n-1} \supsetneq Ke_1 \oplus \cdots \oplus Ke_{n-2} \supsetneq \cdots \supsetneq Ke_1 \supsetneq 0.
            \]

            Hence \(\ell(K^n) = n\).
        \end{enumerate} 
    \end{remark*}

    \begin{definition}
        A module \(N\) is called \underline{indecomposable} if \(N\neq 0\) and \(N\) has no direct summand other than \(0\) and \(N\).

        i.e. We cannot write \(N = A \oplus B\) where \(A, B\) are non-zero submodules. For example, \(K^n\) in the previous example is indecomposable as an \(A\)-module.
    \end{definition}

    \begin{example}
        \begin{enumerate}[label=\roman*)]
            \item Again, suppose \(K\) is a field and \(A = \left\{ n \times n \text{ upper triangular \(K\)-matrices} \right\} \) acting on \(K^n\). Then \(K^n\) is indecomposable as an \(A\)-module.
            \item Any simple module is indecomposable.
            \item Suppose \(K\) is a field, and \(A = K[x]\) acting on \(K^n\) where \(x \mapsto \begin{bmatrix}
                0 & 1 & 0 & 0 & 0  \\
                 & \ddots & \ddots & \ddots & 0 \\
                 &  & \ddots & \ddots & 0 \\
                 &  &  & \ddots & 1 \\
                 &  &  &  & 0 \\
            \end{bmatrix} \) [jordan block]. Then \(K^n\) is indecomposable as an \(A\)-module.
        \end{enumerate} 
    \end{example}

    \begin{proposition}
        If \(M\) is artinian or noetherian, then it is a direct sum of finitely many indecomposable modules.
    \end{proposition}

    \begin{proof}
        Assume \(M\) is artinian. Suppose \(M\neq 0\) and \(M\) is decomposable. Then, \(\exists\) non-zero submodules \(M_1, M_2\) such that \(M = M_1 \oplus M_2\).
        
        Consider \(S(M) = \left\{ 0 \subsetneq N \subsetneq M : N \text{ is a direct summand of } M \right\} \) \(M_1 \in S(M) \implies S(M) \neq \varnothing\).

        \(M\) is artinian. Therefore, \(S(M)\) must have a minimal element \(N_0\). \(N_0\) must be indecomposable, since any non-trivial decomposition of \(N_0\) contraditcs its minimality.

        Write \(M = N_0 \oplus M_0^{\prime}\). 

        Consider \(S^{\prime} = \left\{ M^{\prime} \subset M : \exists M_0 \subset M : M = M_0 \oplus M^{\prime} \text{ and } M_0 \text{ is a finite direct sum of indecomposables}  \right\} \).

        Then, \(M^{\prime}_0 \in S^{\prime}\). Thus, \(S^{\prime} \neq \varnothing\). Therefore \(S^{\prime}\) contains a minimal element which we claim is \(M^{\prime}\).

        Suppose \(M^{\prime} \neq 0\). If \(M^{\prime}\) is indecomposable then \(M = (M_0 \oplus M^{\prime}) \oplus 0\) which contradicts the minimality of \(M^{\prime}\), since \(0 \in  S^{\prime}\).
        
        Hence \(M^{\prime}\) is indecomposable. Therefore, \(S(M^{\prime}) \neq \varnothing\). \(M^{\prime}\) must also be artinian, therefore \(S(M^{\prime})\) has minimal element \(N^{\prime}_0\) which is indecomposable. We write \(M^{\prime} = N^{\prime}_0 \oplus M^{\prime\prime} \implies M = (M_0 \oplus N^{\prime}_0) \oplus M^{\prime\prime}\) where \(M^{\prime\prime}\) is minimal in \(S^{\prime}\), which is a contradiction.

        Noetherian case: HW3.
    \end{proof}

    \begin{lemma}
        Let \(M \in A\text{-Mod}\) and \(g\in \operatorname{End}_A(M)\).

        \begin{enumerate}[label=\roman*)]
            \item If \(g\) is surjective and \(M\) is noetherian \(\implies g\) is injective.
            \item If \(g\) is injective and \(M\) is artinian \(\implies g\) is surjective.
        \end{enumerate} 
    \end{lemma}

    \begin{proof}
        HW3.
    \end{proof}

    \begin{lemma}
        Let \(M\) be of finite length and \(f\in \operatorname{End}_A(M)\). Then \(\exists\) decomposition \(M = U \oplus N\) such that \(f(U) \subset U, f(N) \subset N\) and \(\eval{f}_{U}\) is bijective and \(\eval{f}_{N}\) is nilpotent, \((\eval{f}_{N})^k = 0\). 
    \end{lemma}

    \begin{proof}
        HW3.
    \end{proof}

    \begin{lemma}
        Let \(M\) be of finite length and indecomposable. Then every endomorphism \(f\in C \coloneqq \operatorname{End}_A(M)\) is either bijective or nilpotent.

        Moreover, \(I = \left\{ f\in C : f \text{ is nilpotent} \right\} \) is the unique maximal (\(2\)-sided) ideal of \(C\).
    \end{lemma}

    \begin{proof}
        Suppose \(f\) is not bijective. 1.39 implies that \(M = U \oplus N\) where \(f(U) \subset U, f(N) \subset N\), \(\eval{f}_{U}\) is bijective, \(\eval{f}_{N}\) is nilpotent. Since \(f\) is not bijective, \(M\neq U\). Therefore, \(N \neq 0\). Since \(M\) must be indecomposable, \(U = 0\) and \(M = N\). Therefore, \(f\) must be nilpotent.

        Now we prove that \(I\) must be an ideal.

        \(f\in I, h\in C \implies \ker (h \circ  f) \neq 0 \implies h \circ f \in I\).

        \(f\in I, h\in C \implies \operatorname{im} (f \circ h) \subsetneq M \implies f \circ h\) is not bijective \(\implies f \circ h \in I\). 

        We need to show \(I\) is closed under addition. Suppose \(f,g \in I\) but \(f+g\notin I\). Then \(f+g\) must be invertible.

        \(h(f+g)=1 \implies hg = 1-hf\). Since \(hf\) is nilpotent it follows that \(hg\) is invertible, which is a contradiction.
    \end{proof}

    \begin{theorem}
        [Krull-Remak-Schmidt] Let \(M\) be a module of finite length. Suppose \(M = N_1 \oplus \cdots \oplus N_m = N^{\prime}_1 \oplus \cdots \oplus N^{\prime}_n\) with indecomposable submodules \(N_i, N_j^{\prime}\). Then \(m = n\) and \(\exists \sigma \in \mathcal{S}_n\) such that \(\forall 1 \leq i \leq n : N_i^{\prime} \cong N_{\sigma(i)}\).
    \end{theorem}

    \begin{proof}
        By induction on \(\ell(M)\). If \(\ell(M) = 1\) there is nothing to show. Let \(\ell(M) > 1\) and suppose the sttement is true for all modules of length \(< \ell(M)\).

        Let \(\iota_j : N_j \hookrightarrow M\) be the inclusion and \(\pi_j: M \to N_j\) the projection onto \(N_j\).

        Let \(p_j = \iota_j \circ \pi_j \in \operatorname{End}_A(M)\). Then \(p_j\) is an idempotent, \(p_j \circ p_j = p_j\) and \(p_j \circ p_i = 0\) if \(i\neq j\). Furthermore \(p_1 + \cdots + p_m = \operatorname{id}_{M}\).

        We can do the same for the other decomposition: we get \(p_j^{\prime}\) where \(p^{\prime}_1 + \cdots + p^{\prime}_n = \operatorname{id}_{M}\) etc.
        
        Consider: \(\operatorname{End}_A(N_1)\ni f_j = \pi_1 \circ p_j^{\prime} \circ \iota_1: N_1 \hookrightarrow M \xrightarrow{\pi_j^{\prime}} N^{\prime}_j \xhookrightarrow{\iota^{\prime}_j} M \to N_1\).
        
        \(f_1 + f_2 + \cdots + f_n = \pi_1 \circ  \underbrace{(p^{\prime}_1 + \cdots +p^{\prime}_n)}_{\operatorname{id}_{M}} \circ \iota_1 = \operatorname{id}_{N_1}\).        

        1.39 implies that \(\exists 1 \leq j \leq n\) such that \(f_j\) is bijective. After renumbering, we may assume that \(f_1\) is bijective.

        Consider \(g=p^{\prime}_1 \circ p_1: M \to M\). Note that \(\pi_1 \circ g = f_1 \circ \pi_1\).


        Set \(h = g + p_2 + \cdots + p_m : M \to M\). Then \(p_1 \circ h = p_1 \circ  g = \iota_1 \circ \pi_1 \circ g = \iota_1 \circ f_1 \circ \pi_1\). We want to show that \(h\) is bijective.

        Suppose \(x\in M\) such that \(h(x) = 0\).

        \(\implies 0 = p_1(h(x)) = p_1(g(x)) = \iota_1(f_1(\pi_1(x))) = 0\). Since \(\iota_1\) and \(f_1\) are injective, it follows that \(\pi_1(x) = 0\). Therefore \(x\in N_2 \oplus \cdots \oplus N_m \eqqcolon M^{\prime}\). Since \(g = p^{\prime}_1 \circ p_1\) it follows that \(g(x) = 0\).

        \(0 = h(x) = g(x) + p_2(x) + \cdots + p_m(x) = x\) where \(g(x) = 0\) and \(p_j(x) \in N_j\). Therefore \(h\) is injective.

        It follows from 1.37 that \(h\) is bijective.

        Since \(h\) is an automorphism and \(\eval{h}_{N_k} = \operatorname{id}_{N_k}\) for \(k \geq 2\) it follows that \(M = N^{\prime}_1 \oplus N_2 \oplus \cdots \oplus N_m\). By quotienting out \(N^{\prime}_1\) it follows that \(N_2 \oplus \cdots \oplus N_m \cong N^{\prime}_2 \oplus \cdots \oplus N^{\prime}_n\). The theorem follows from induction.

    \end{proof}
    
    \section*{Tuesday, 2/3/2026}
    
    \section{Wedderburn Theory}

    Recall: An algebra \(A\) is called \underline{simple} if \(A\) is non-zero and it has no ideals other than \(0\) and \(A\).

    Recall that by ideal we mean ideals that are both left and right ideals. There can be left ideals that are not right ideals and vice versa. Meaning, an algebra \(A\) being simple doesn't necessarily imply that the module \(A_\ell\) is simple.

    \begin{proposition}
        Let \(A\) be simple. TFAE:

        \begin{enumerate}[label=\roman*)]
            \item \(A\) is semisimple.
            \item \(A\) is artinian.
            \item \(A\) possesses a minimal left ideal \(N\). 
        \end{enumerate} 
    \end{proposition}

    Minimal ideals are by convention non-zero. This forces \(A\) to be non-zero from iii as well.

    Recall by saying \(A\) is noetherian/artinian we mean \(A_\ell\) is noetherian/artinian.

    \begin{proof}
        i \(\implies\) ii: Corollary 1.33 (because \(A_\ell\) is semisimple and finitely generated, hence it has finite length).

        ii \(\implies\) iii: Trivial.

        iii \(\implies\) i: Let \(0 \neq N \subset A\) be the minimal left ideal. Then, \(0 \neq NA = \sum_{a\in A} Na\) is a \(2\)-sided ideal. Since \(A\) is simple, \(NA = A\).

        \(N\) minimal left ideal \(\implies N\) is simple as an \(A\)-module. \(Na\) is the image of \(N\) under the \(A\)-module map \(N \to Na, x \mapsto xa\). If \(Na\) is non-zero, it is the non-zero image of a simple module, which implies \(Na\) is simple. Then, \(NA\) is a sum of simple modules: \(A = NA = \sum_{Na\neq 0} Na\) is a sum of simple modules. Therefore, \(1.16 \implies A\) must be semisimple.
    \end{proof}

    \begin{corollary}
        Let \(A\) be simple and semisimple and \(N \subset A\) a minimal left ideal. Then, \(A_\ell \underset{A}{\cong} N^{\oplus m}\) for some \(m > 0\), and \(A_\ell\) is hence isotypic.
    \end{corollary}

    \begin{proof}
        2.1 implies the existence of a minimal ideal \(N\). Then, \(A_\ell = \sum_{a\in A, Na\neq 0} Na\) where \(Na\) are simple. By 1.17, \(A_\ell \cong \bigoplus_{i\in I} Na_i\) for some subset \(\{ a_i : i \in I \} \) of \(A\). WLOG we assume that \(Na_i \neq 0\) for all \(i\in I\).

        Then, the identity \(1_A = \sum_{j\in J \subset I} n_j a_j\) for some finite \(J \subset I\) and \(n_i \in N\). Therefore, \(\forall a\in A: a = \sum_{j\in J} a n_j a_j\). Note that \(a n_j a_j \in Na_j\) since \(N\) is a left ideal. Therefore, \(A = \bigoplus_{i\in J} Na_i\). Therefore \(J = I\). Furthermore, for each \(i\in I\), \(Na_i \cong N\).
        
        Therefore, \(A_\ell \cong \bigoplus_{i\in I} Na_i \cong N^{\oplus \vert I \vert}\).

    \end{proof}

    \begin{proposition}
        Let \(A\) be a non-zero semisimple algebra. TFAE:

        \begin{enumerate}[label=\roman*)]
            \item \(A\) is simple.
            \item \(A_\ell\) is isotypic.
            \item \(\vert \mathcal{T} (A) \vert = 1\). Recall that \(\mathcal{T} (A)\) is the set of isomorphism classes of simple \(A\)-modules.
        \end{enumerate} 
    \end{proposition}

    \begin{proof}
        i \(\implies\) ii follows from 2.2.

        ii \(\implies\) iii follows from 1.18: every simple module is isomorphic to a submodule of \(A_\ell\).
        
        Recall 1.24 which says that minimal ideals are precisely the isotypic components of \(A_\ell\). in fact, \(A_\ell = \bigoplus_{\tau \in \mathcal{T}(A)} (A_\ell)_\tau\).
        
        If \(N\) is simple of type \(\tau \implies (A_\ell)_{\tau} \cong \bigoplus_{i\in I_\tau} N\).
        
        If \(N\) is simple \(A_\ell / L \cong N \to N\) appears as a direct summand in \(A_\ell = \bigoplus_{\tau \in \mathcal{T}(A)} (A_\ell)_{\tau}\). \((A_\ell)\tau \neq 0\) for all \(\tau \in \mathcal{T}(A)\).
        
        iii \(\implies\) ii is trivial.

        ii \(\implies\) i follows from 1.24.

        Summarizing,

        \[
            \begin{tikzcd}[column sep = huge]
                \text{i} \ar[r,Rightarrow, bend left, "2.2"] & \text{ii} \ar[r, Rightarrow, bend left, "1.18"] \ar[l, Rightarrow, bend left, "1.24"] & \text{iii} \ar[l, Rightarrow, bend left, "\text{trivial}"]
            \end{tikzcd}
        \]
    \end{proof}

    \begin{proposition}
        If \(D\) is a division algebra, then \(M_n (D)\) is a simple artinian algebra for any \(n \geq 1\).
    \end{proposition}

    \begin{proof}
        Set \(V = (D^o)^{\oplus n} \implies A \coloneqq \operatorname{End}_{D^o}(V) \cong M_n((D^o)^o) = M_n(D)\).

        By 1.13: \(A_\ell \overset{(\ast)}{=} \underbrace{V \oplus \cdots \oplus V}_{n \text{ copies}}\) and \(V\) is simple as \(A\)-module (1.13). Then \(A_\ell\) is isotypic and seimisimple because of \(\ast\). Then by 2.3 \(A\) is simple. 
    \end{proof}

    Recall: \(Z(A) \coloneqq\) center of \(A\).

    \begin{proposition}
        \(A\) simple \(\implies Z(A)\) is a field.
    \end{proposition}

    \begin{proof}
        Pick non-zero element of the center \(a\in Z(A) \setminus \{ 0 \}\). Note that \(Aa=aA\) must be a non-zero two-sided ideal, and since \(A\) is simple it follows that \(aA = A\). \(aA = A \implies \exists b\in A\) such that \(ab (=ba) = 1\). Therefore \(a\) must have a two-sided inverse.

        Given \(c\in A: (cb - bc) a = c(ba) - (ba) c = c1 - 1c = c-c = 0\). Multiplying by \(b\) on the right, it follows that \((cb - bc) ab = 0 \implies cb - bc = 0  \implies cb = bc\) for all \(c\in A\). Therefore \(b\in Z(A)\).
    \end{proof}


    \begin{theorem}
        A non-zero semisimple algebra \(A\) has only finitely many distinct minimal ideals \(A_1, \cdots , A_n\). Each \(A_i\) is a unital algebra in its own right with the induced addition and multiplication from \(A\). Note that the identity in \(A_i\) is not equal to the identity in \(A\) if \(n > 1\).

        Moreover, \(A = A_1 \times \cdots \times A_n\) is the product of the algebras \(A_1 , \cdots , A_n\) (with componentwise addition and multiplication), and each \(A_i\) is simple and artinian.

        Conversely, if \(A_1 , \cdots , A_n\) are simple artinian algebras, then \(A \coloneqq A_1 \times \cdots \times A_n\) is semisimple and \(A_1, \cdots , A_n\) are precisely the minimal ideals of \(A\).
    \end{theorem}

    Note that direct sum and product carry the same meaning in this case. We generally use direct sums when we're thinking about the object as a module, and products when we're thinking about the object as a ring.

    \begin{proof}
        1.24 \(\implies A = A_1 \oplus \cdots \oplus A_n\) where \(A_1, \cdots , A_n\) are the isotypic components of \(A_\ell\). These are precisely the (2-sided) ideals.

        We also have: \(A_i A_j \subset A_1 \cap A_j = 0\) for \(i\neq j\).

        Write \(1_A = e_1 + \cdots + e_n\) with \(e_i \in A_i\). \(1_A = 1_A \cdot 1_A \implies 1_A = e_1^2 + \cdots + e_n^2\). Taking projections it follows that \(\forall 1 \leq i \leq n: e_i^2 = e_i\).
        
        \(\forall a\in A_i, a = a\cdot 1_A = ae_1 + \cdots + ae_i + \cdots + ae_n = ae_i = e_i a\).
        
        Therefore, \(e_i\) is the identity element in \(A_i\).

        If \(0\neq I \subset A_i\) is an ideal, then \(I\) is an ideal in \(A\). However, since \(A_i\) is minimal, it follows that \(I = A_i\). Therefore \(A_i\) must be simple.

        If \(0 \neq I \subset A_i\) is a left ideal, then \(I\) is a left ideal in \(A\). Since \(A\) is a semisimple algebra \(A\) must be artinian. Thus \(A_i\) must also be artinian.

        The converse is easy to check (HW4)
    \end{proof}

    \begin{corollary}
        Let \(A = A_1 \times \cdots \times A_n\) be a semisimple algebra with simple algebras \(A_1, \cdots , A_n\). Then,
        
        \[
            Z(A) = Z(A_1) \times \cdots \times Z(A_n)
        \]

        is a product of fields, by 2.5. IN particular, \(Z(A)\) is a field if and only if \(A\) is simple.
    \end{corollary}

    \begin{corollary}
        A commutative semisimple algebra \(A\) is a product of finitely many fields: \(A = K_1 \times \cdots \times K_n\).

        These fields are uniquely determined as subsets of \(A\). Namely, they are the minimal ideals of \(A\).
    \end{corollary}

    \begin{remark}
        A commutative artinian algebra \(A\) is semisimple if and only if its nilradical \(\{ a\in A \mid \exists n > 0 : a^n = 0 \} \) is zero.

        \(\implies\) follows from 2.8.

        \(\impliedby\) uses the theory of redical at the end of chapter 28. Direct proof on HW4.
    \end{remark}

    \begin{corollary}
        Let \(K\) be a field. A finite dimensional commutative \(K\)-algebra having no non-zero nilpotent elements is a product of finitely many field extensions \(K_i / K\) with \([K_i : K] < \infty\).
    \end{corollary}

    \begin{proof}
        By the preceding remark, \(A\) is semisimple. By 2.8, it is a product of fields \(K_1, \cdots , K_n\), all of which have finite degree over \(K\).
    \end{proof}

    Now we state a key theorem of this course.

    \begin{theorem}
        [Wedderburn's Theorem]

        An artinian algebra \(A\) is simple if and only if it is isomorphic (as rings) to a matrix algebra \(M_n(D)\) for a division algebra \(D\) and \(n > 0\):

        \[
            A \cong M_n(D)
        \]

        \(D\) and \(n\) are uniquely determined by \(A\).
    \end{theorem}

    \begin{proof}
        By the remark, \(A\) is semisimple. By 2.8 it is a product of fields \(K_1, \cdots , K_n\), all of which have finite degree \(/ K\).
    \end{proof}

    \(A\) semisimple \(\implies A \cong M_{n_1}(D_1) \times \cdots \times M_{n_r}(D_r)\).

    \([D_1] \cdot [D_2] = [D_1 \otimes_K D_2] = [M_n(D_3)] = [D_3]\). 
    
    \section*{Thursday, 2/5/2026}
    
    \begin{remark*}
        Prop 2.4: \(M_n(D)\) is simple.
    \end{remark*}

    \begin{notation*}
        If \(K\) is a (skew) field then \([V:K] \coloneqq \dim_K(V)\).
    \end{notation*}

    \begin{theorem}
        Let \(A\) be a simple artinian algebra, and let \(N\) be a simple \(A\)-module (unique upto isomorphism). Set \(D = \operatorname{End}_A(N)\) which is a division algebra.

        \begin{enumerate}[label=\roman*)]
            \item The canonical map \(A \to \operatorname{End}_D(N), a \mapsto a_N, a_N(x) = ax\), is an isomorphism, and \(r \coloneqq \dim_D(N)\) is finite and equal to \(\ell_A(A_\ell)\).
            
            Therefore, \(A \cong M_r(D^o)\). \(Z(A)\) is a field, isomorphic to \(Z(D)\).

            \item If \(M\) is a non-zero f.g. \(A\)-module, then \(B = \operatorname{End}_A(M)\) is simple and artinian, and \(B\cong M_n(D)\), with \(D\) as in (i)and \(n = \ell_A(M)\). If \(K\) is a field and \(A\) is a \(K\)-algebra, then,
            
            \[
                [M:K]^2 = [A:K][B:K]
            \]

            and all these dimensions are finite provided one of them is finite.
        \end{enumerate} 
    \end{theorem}

    \begin{proof}
        Step 1: 2.1 implies \(A\) is semisimple. 2.3 \(\implies \vert \mathcal{T} (A) = 1 \vert.\) Let \(N\) be a simple \(A\)-module, \(M\) a f.g. \(A\)-module. 1.19 \(\implies M\) is semisimple. \(\overset{1.32}{\implies} M \cong N^{\oplus n}\). \(n = \ell_A(M)\).

        \(1.7 \implies B \coloneqq \operatorname{End}_A(M) \cong \operatorname{End}_A(N^n) \cong M_n(\operatorname{End}_A(N)) \overset{def}{=} M_n(D)\).

        Thus \(B\) is simple and artinian.

        Step 2: Apply Step 1 with \(M = A_\ell \cong N^{\oplus m}, m = \ell_A(A_\ell)\).

        Step 1 \(\implies \operatorname{End}_A(A_\ell \cong M_m(\operatorname{End}_A(N))) = M_m(D)\).

        Proposition 1.5 (HW1): \(\operatorname{End}_A(A_\ell) = A^o\).

        \(\implies A\cong \operatorname{End}_A(A_\ell)^o \cong M_m(D)^o \overset{HW1}{\cong} M_m(D^o)\).
        
        Step 3: It remains to show: \(\boxed{m (= \ell_A(A_\ell)) = \dim_D (N) \eqqcolon r.}\)

        Fix an isomorphism of \(A\)-modules \(\varphi : A_\ell \xrightarrow{\cong}N^{\oplus m}\).

        Consider the diagram:

        \[
            \begin{tikzcd}
                A \ar[r,"a \mapsto a_N"] \ar[d,"\alpha","1.5\implies\cong"'] & \operatorname{End}_D(N) \ar[d,"\beta", "1.7 \implies \cong"'] \\ \operatorname{End}_{A^o}(A^o_\ell) = \operatorname{End}_{\operatorname{End}_A(A_\ell)}(A_\ell^o) \ar[r,"\cong","\varphi^{\ast}"'] & \operatorname{End}_{\operatorname{End}_A(N^m)}(\underbrace{N^m}_{=N^{\prime} \text{ in 1.7}})
            \end{tikzcd}
        \]

        \[
            \varphi^{\ast} (\tau)(\underbrace{x}_{\in N^{\prime} = N^m}) = \varphi (\tau (\varphi ^{-1} (x))).
        \]

        Easy to check: this diagram is commutative.

        Upshot: the canonical map \(A \to \operatorname{End}_D(N)\) is an isomorphism!

        \(\implies \operatorname{End}_D(N)\) is artinian.

        \(\implies \operatorname{End}_D(N)\) is artinian \(\underset{\text{easy (HW4)}}{\implies} [N:D] = \dim_D(N) < \infty\).
        
        \(\implies N \underset{D}{\cong} D^{\oplus r}\).

        1.13 \(\implies \operatorname{End}_D(N)_\ell \underset{\text{as } \operatorname{End}_D(N) \text{-modules} }{\cong} N^r \).

        In the other hand, \(\operatorname{End}_D(N)_\ell \cong M_n(D^o) \cong \operatorname{End}_D(D)\).
        
        Therefore, \(r = \dim_D(N) = \ell_A(A_\ell) = m\).
        
        Consider again \(A \xrightarrow{\cong} \operatorname{End}_D(N) \cong M_r(D^o)\).

        \(\implies Z(A) \xrightarrow{\cong} Z(M_r(D^o)) \underset{\text{easy (HW4)}}{=} Z(D^o) = Z(D)\).
        
        Step 4: Let \(M\in A\text{-Mod}^{\text{f.g.}}\). Assume \(A\) is a \(K\)-algebra. Write \(M = N^n\).

        Then, \([M:K] = n[N:K] = n[N:D][D:K] = nr [D:K]\).

        i \(\implies [A:K] = r^2 [D:K]\).

        Step 1 \(\implies [B:K] = n^2[D:K]\).

        This shows the formula.

    \end{proof}

    \begin{theorem}
        Suppose \(D_1, D_2\) are division algebras and \(M_r(D_1) \cong M_s(D_2)\) as rings. Then \(r=s\) and \(D_1 \cong D_2\).
    \end{theorem}

    \begin{proof}
        Step 1: Suppose \(N, N^{\prime}\) are \(A\)-modules over \(A\) (any algebras). Then any isomorphism f\(: N \to N^{\prime}\) of \(A\)-modules induces an isomorphism of rings \(\operatorname{End}_A(N) \to \operatorname{End}_A(N^{\prime})\) given by \(\tau \mapsto f \circ \tau \circ f ^{-1}\).

        Step 2: If \(g: A \to A^{\prime}\) is an isomorphism of simple artinian algebras, and \(N\in A\text{-Mod}\) and \(N^{\prime} \in A^{\prime}\text{-Mod}\) are simple, then \(a \cdot x^{\prime} \eqqcolon g(a) \cdot x^{\prime}\) gives \(N^{\prime}\) the structure of an \(A\)-module. Since \(g\) is an isomorphism, \(N^{\prime}\) becomes a simple \(A\)-module. \(2.3 \implies N^{\prime} \cong N\) as \(A\)-modules. 

        Step 1 \(\implies \operatorname{End}_A(N) \cong \operatorname{End}_A(N^{\prime}) = \operatorname{End}_{A^{\prime}}(N^{\prime})\) immediately.

        Step 3: \(D\) division algebra \(\underset{1.13}{\implies} D^r\) is a simple module over \(A\coloneqq \operatorname{End}_D(D^r) (\cong M_n(D^o))\) and \(\operatorname{End}_{M_r(D^o)}(D^r) \underset{1.8}{\cong} D\).

        Now suppose \(M_r(D_1) \cong M_s(D_2)\).

        HW1 \(\implies M_r(D_1^o) \underset{\text{HW1}}{\cong} M_n(D_1)^o \cong M_s(D_2)^o \underset{\text{HW1}}{\cong} M_s(D_2^o)\).
        
        \(N \coloneqq D_1^r\) is a simple \(M_r(D_1^o)\)-module.

        \(N^{\prime} \coloneqq D_2^s\) is a simple \(M_s(D_2^o)\)-module.

        Step 2\(\implies D_1 \underset{(\ast)}{\cong} \operatorname{End}_{M_r(D_1^o)} (N) \cong \operatorname{End}_{M_s(D_2^o)}(N^{\prime}) \underset{(\ast)}{\cong} D_2 \).

        Step 4: If \(A \coloneqq M_r(D) \underset{\text{rings}}{\cong} M_s(D) \eqqcolon B\) then \(r = \ell_A(A_\ell) = \ell_B(B_\ell) = s\). 
    \end{proof}

    \begin{proof}
        [Proof of Wedderburn's Theorem] Combine 2.12 and 2.13.
    \end{proof}

    \begin{corollary}
        \begin{enumerate}[label=\roman*)]
            \item A simple algebra \(A\) is artinian if and only if \(A^o\) is artinian, in which case \(\ell_A(A_\ell) = \ell_{A^o}(A_\ell^o)\).
            \item \(A\) is semisimple if and only if \(A^o\) is semisimple, in which case \(\ell_A(A_\ell) = \ell_{A^o}(A^o_\ell)\). 
        \end{enumerate} 
    \end{corollary}

    \begin{proof}
        \begin{enumerate}[label=\roman*)]
            \item \(A\) simple artinian \(\underset{2.13}{\implies} A \cong M_n(D) \implies A^o \cong M_n(D^o)\).
            \item follows from 2.6 and (i).
        \end{enumerate} 
    \end{proof}

    \begin{corollary}
        A unital associative algebra \(A\) is semisimple if and only if \(A \cong M_{n_1}(D_1) \times \cdots \times M_{N_r}(D_r)\) where \(D_1 , \cdots , D_r\) are division algebras and \(n_1, \cdots , n_r\) and the isomorphism classes of \(D_r, \cdots , D_r\) are uniquely determined upto permutation. 
    \end{corollary}

    \section*{Tuesday, 2/10/2026}
    
    An immediate consequence of corollary 2.15 is as follows:

    \begin{theorem}
        [HW4] A finite dimensional \(K\)-algebra \(A\) is simple if and only if it is isomorphic to \(M_n(D)\) [as \(K\)-algebras] where \(D\) is a finite dimensional division algebra over \(K\).

        One also has \(Z(A) \underset{\text{HW}}{=} Z(D)\).
    \end{theorem}

    Then, \(\dim_K(A) = n^2 \cdot \dim_K(D)\).

    Another easy consequence:

    \begin{theorem}
        If \(K\) is algebraically closed and \(A\) is a finite dimensional simple \(K\)-algebra, then \(A \cong M_n(K)\) as \(K\)-algebras for some \(n > 0\).
    \end{theorem}

    \begin{corollary}
        Let \(A\) be an algebra over an algebraically closed field \(K\), and let \(M\) be a simple \(A\)-module satisfying \([M : K] < \infty\).

        Then, the canonical map \(K \to \operatorname{End}_A(M), \lambda \mapsto \lambda \cdot \operatorname{id}_{M}\) is an isomorphism, and \(A \to \operatorname{End}_K(M), a \mapsto a_M\) is surjective.

        If \(A\) is commutative, we have \([M:K] = 1\).
    \end{corollary}

    \begin{proof}
        HW5.
    \end{proof}

    \begin{definition}
        Two simple artinian \(R\)-algebras \(A, B\) are called \underline{similar} if there exists an \(R\)-division algebra \(D\) and \(R\)-algebra isomorphisms \(A \cong M_n(D)\) and \(B\cong M_m(D)\). 
    \end{definition}

    \begin{notation*}
        \(A \underset{R}{\sim} B \iff A\) and \(B\) are similar.
    \end{notation*}

    \begin{remark*}
        \begin{enumerate}[label=\arabic*)]
            \item \(A \underset{R}{\cong} B \implies A \underset{R}{\sim} B\).
            \item \(A \underset{R}{\sim} B\) and \(\ell_A(A) = \ell_B(B) \implies A \underset{R}{\cong} B\).
            \item \(\underset{R}{\sim}\) is an equivalence relation.
            \item \(A \underset{R}{\sim} B \implies Z(A) = Z(B)\).
        \end{enumerate} 
    \end{remark*}

    General convention: \(K\) is always considered a field. \(K\)-Alg \(=\) category of unital associative \(K\)-algebras.

    \begin{definition}
        Let \(A\in K\text{-Alg}\). \(A\) is called:

        \begin{enumerate}[label=\roman*)]
            \item \underline{central} if \(Z(A) = K\).
            \item \underline{central simple} if \(A\) is central, simple, and \([A:K] < \infty\).
            \item \(\operatorname{CSA}(K) = \) category of central simple \(K\)-algebras with \(K\)-algebra homomorphisms as morphisms.
        \end{enumerate} 
    \end{definition}

    \begin{remark*}
        Let \(A\in K\text{-alg}\)

        \(A \in \operatorname{CSA}(K) \iff A \cong M_n(D)\) with \(D\in \operatorname{CSA}(K)\) \([D:K] < \infty , Z(D) = K\).
    \end{remark*}

    \begin{definition}
        A set of \underline{similarity} classes of central simple \(K\)-algebras is called the \underline{Brauer Group} of \(K\), denoted \(\operatorname{Br}(K)\).
    \end{definition}

    We thus have \(\operatorname{Br}(K) =\) set of \(K\)-isomorphism classes of central simple \(K\)-division algebras.

    \begin{remark*}

        \begin{enumerate}[label=\arabic*)]
            \item  Richard Brower, 1901-1977. Kentrucky, Toronoto, Michigan, Harvard. Visited E. Artin in Bloomington in 1941.
            \item \(\operatorname{Br}(K)\) is indeed a set: \(A \hookrightarrow \operatorname{End}_K(A) \cong M_n(K), n = [A:K]\).
            \item We write \([A]\) for an object in \(\operatorname{Br}(K)\) represented by \(A\in \operatorname{CSA}(K)\).
        \end{enumerate} 

    \end{remark*}

    \underline{Towards the group structures of \(\operatorname{Br}(K)\):}

    \underline{Convention}: \(\otimes = \otimes_K\).

    \begin{proposition}
        Let \(A\) be a \(K\)-algebra.

        \begin{enumerate}[label=\roman*)]
            \item Setting \((a \otimes b) \cdot x = axb, a,b,x\in A\) gives \(A\) the structure of a module over \(A \underset{\otimes}{K} A^0\).
            \item \(A\) is a simple \(K\)-algebra \(\iff A\) is a simple \(A \otimes A^o\)-module.
            \item \(\operatorname{End}_{A \otimes_K A^o}(A) = Z(A)\).
            \item If \(B \subset A\) is a \(K\)-subalgebra, then \(B \otimes_K A^o \subset A \otimes_K A^o\) is a subalgebra, and:
            
            \[
                \operatorname{End}_{B \otimes_K A^o}(A) = Z_A(B) = \left\{ a\in A : \forall b\in B: ab = ba \right\}\;.
            \]

            is called the \underline{centralizer of \(B\) in \(A\)}.
        \end{enumerate} 
    \end{proposition}

    \begin{proof}
        \begin{enumerate}[label=\roman*)]
            \item Immediate.
            \item A \(A \otimes A^o\)-submodule of \(A\) is an ideal and vice versa.
            \item is a consequence of iv.
            \item Consier the map \(Z_A(B) \to \operatorname{End}_{B \otimes A^o}(A)\) given by \(a \mapsto \varphi_a, \varphi_a(x) = ax\). The map in the other direction is given by \(\varphi(1) \mapsfrom \varphi\).  
        \end{enumerate} 
    \end{proof}

    \begin{proposition}
        Let \(A,B \in K\)-alg, \(A^{\prime} \subset A, B^{\prime} \subset B\) subalgebras.

        Then the map \(Z_A(A^{\prime}) \otimes_K Z_B(B^{\prime}) \to Z_{A \otimes B} (A^{\prime} \otimes B^{\prime})\) given by \(a \otimes b \mapsto a \otimes b\) where \(a\in Z_A(A^{\prime}), b\in Z_B(B^{\prime})\) is an isomorphism of \(K\)-algebras.
    \end{proposition}

    \begin{proof}
        Let \((a_i)_{i\in I}\) be a \(K\)-basis of \(A\), and \(z\in Z_{A \otimes B}(A^{\prime} \otimes B^{\prime})\).

        Write \(z = \sum_{i} a_i \otimes b_i\) with uniquely determined \(b_i\).

        Given \(b\in B^{\prime}\), we find that \(\sum_{i} a_i \otimes (bb_i) = (1 \otimes b) z = z (1 \otimes b) = \sum_{i} a_i \otimes (b_i b)\).
        
        By uniqueness, \(\forall i, bb_i = b_i b\). Therefore, \(\forall i, b_i \in Z_B(B^{\prime}).\) It follows that \(z \in A \otimes Z_B(B^{\prime})\).

        We can apply the same argument with a \(K\)-basis of \(B\) implies that \(Z \in Z_A(A^{\prime}) \otimes_K B\).

        Then, \(z\in A \otimes_K Z_B(B^{\prime}) \cap Z_A(A^{\prime}) \otimes_K B = Z_A(A^{\prime}) \otimes_K Z_B(B^{\prime})\).
    \end{proof}

    \begin{remark*}
        The tensor product of simple algebras need not be simple.

        \(\mathbb{C}\) is a simple \(\mathbb{R}\)-algebra. However, \(\mathbb{C} \otimes_\mathbb{R} \mathbb{C} \cong \mathbb{C} \times \mathbb{C}\) as \(\mathbb{R}\)-algebras.
    \end{remark*}

    \begin{theorem}
        Let \(A,B \in K\)-Alg be simple and suppose at least one of them is central.

        Then, \(A \otimes_K B\) is simple.
    \end{theorem}

    \begin{proof}
        WLOG suppose \(Z(A) = K\). Let \(T \subset A \otimes B\) be any (two-sided by default) ideal.

        \underline{Claim}: \(T = A \otimes I\), where \(I = T \cap B\) [where we think of this \(B\) as \(K \otimes_K B \subset A \otimes_K B\).]

        \underline{Proof of Claim}: Let \(t\in T\) and write \(t = \sum_{i=1}^n x_i \otimes y_i\), where \(\{ x_1, \cdots , x_n \} \) is linearly independent over \(K\).

        Set \(\mathscr{M} = A, \mathscr{A} = A \otimes_K A^o\) and \(\mathscr{C} = \operatorname{End}_{\mathscr{A}} (\mathscr{M}) = \operatorname{End}_{A \otimes A^o}(A) \overset{2.19}{=} Z(A)= K\).

        Apply Jacobson's Density Theorem to \(\mathscr{A} = A \otimes A^o \to \operatorname{End}_{\mathscr{C}}(\mathscr{M}) = \operatorname{End}_K (A)\) and choose \(f_i \in \operatorname{End}_K(A)\) such that \(f_i(x_j) = \delta_{ij} 1_A\), which exists since \(x_1, \cdots , x_n\) are linearly independent.

        1.26 \(\implies \exists \alpha \in \mathscr{A} = A \otimes A^o\) such that \(\alpha_i \cdot x_j = \delta_{ij}1_A\).

        Consider \(A \otimes_K B\) as a module over \(A \otimes A^o\) by \((a_1 \otimes a_2) (\underset{\in A}{x} \otimes \underset{\in B}{y})= (a_1 x a_2) \otimes y\).
        
        Note that \((a_1 \otimes a_2)(x \otimes y) = (a_1 \otimes 1) (x \otimes y) (a_2 \otimes 1)\). Then, if we have a two-sided ideal of \(A \otimes_K B\), it would be stable under the action of \(A \otimes A^o\).

        \(\implies T\) is an \(A \otimes A^o\)-submodule of \(A \otimes B\).
        
        Therefore, \(\alpha_i \cdot t = \sum_{j} (\alpha_i \cdot x_j) \otimes y_j = 1_A \otimes y_i \in T \cap (K \otimes_K B) = T \cap B = I\).

        It follows that \(A \otimes I \supset T \supset A \cdot \underset{\subset T}{I}  = (A \otimes K) \cdot (K \otimes_K I)\). 

        Therefore \(T = A \otimes I\).

    \end{proof}

    \begin{remark*}
        \(I = T\cap B\) is a \(2\)-sided ideal in \(B\).

        \(B\) simple \(\implies I = 0\) or \(I = B \implies T = A \otimes 0 = 0\) or \(T = A \otimes B\). 
    \end{remark*}

    \begin{proposition}
        \(A, B \in K\)-Alg.

        \begin{enumerate}[label=\roman*)]
            \item \(A \otimes B\) simple \(\implies A\) \underline{and} \(B\) are simple.
            \item \(A \otimes B\) artinian \(\implies A\) and \(B\) are artinian.
            \item If \([A : K] < \infty\) and \(B\) is artinian, then \(A \otimes B\) is artinian. 
        \end{enumerate} 
    \end{proposition}

    \begin{proof}
        i and ii are easy.

        iii) Let \([A:K] = n\). Since \(A\) is finite dimensional, as a left \(B\)-module, \(A \otimes_K B \cong K^n \otimes_K B\). But this is isomorphic to \(B^n\) as a left \(B\)-module, which implies that it must be artinian by corollary 1.31. 
    \end{proof}
    
    \section*{Thursday, 2/12/2026}

    Rings \(R\) and \(S\) are called \underline{Monta equivalent} if \(R\)-mod and \(S\)-mod are equivalent as additive categores.

    Action \(D \curvearrowright V\) gives us a corresponding action \(M_n(D) \curvearrowright V^n\).
    
    Functor \(D\)-Mod \(\to M_n(D)\)-Mod, \(V \rightsquigarrow V^{\oplus n}\) is an equivalence of additive categories.
    
    If \(A, B \in \operatorname{CSA}(K)\) and are \(K\)-Monta equivalent theb \(A\) and \(B\) are similar as \(K\)-algebras.

    \[
        A \otimes_K B \underset{B\text{-module}}{\cong} K^n \otimes_K B \cong B^{\oplus n}\;.
    \]
    
    \begin{proposition}
        \(A, B \in K\)-Alg simple artinian. If at least one of them is f.d. over \(K\) and at least one of them is central, then \(C = A \otimes_K B\) is simple and artinian. 
    \end{proposition}

    \begin{proof}
        2.21 \(\implies C\) is simple.

        2.22(iii) \(\implies C\) is artinian.
    \end{proof}

    \begin{theorem}
        \(A, B \in \operatorname{CSA}(K) \implies C\coloneqq A \otimes_K B \in \operatorname{CSA}(K)\).
    \end{theorem}

    \begin{proof}
        \(A, B\) simple (and artinian) \(\implies C\) is simple.

        \([C:K] = [A:K][B:K] < \infty\).

        \(Z(C) \underset{2.20}{=} Z(A) \otimes_K Z(B) = K \otimes_K K \xrightarrow{\cong} K\) by \(a \otimes b \mapsto ab\).
    \end{proof}

    \begin{proposition}
        \(A \in \operatorname{CSA}(K) \implies\) the canonical map \(A \otimes_K A^o \to \operatorname{End}_K (A), a \otimes b \mapsto [x \mapsto axb]\), is an isomorphism of \(K\)-algebras. Hence, \(A \otimes_K A^{\prime} \cong M_n(K), n = [A:K]\). In particular, \([A \otimes_K A^o] = [K]\).
    \end{proposition}

    \begin{proof}
        2.24 [\(A^o \in \operatorname{CSA}(K)\)]  implies \(A \otimes_K A^o\) simple \(\implies \ker (\gamma)\) is a \(2\)-sided ideal and \(\gamma \neq 0 \implies \ker (\gamma) = 0\).

        Comparing dimensions, we see \(\gamma\) is an isomorphism.
    \end{proof}

    \begin{proposition}[Definition of Base Change]
        Let \(A\in K\)-Alg and \(L / K\) a field extension. Then \(A_L = L \otimes_K A\) is an \(L\)-algebra, called the \underline{base change} of \(A\) to \(L\).


        \begin{enumerate}[label=\roman*)]
            \item \([A_L : L] = [A:K]\).
            \item \(Z(A) = K \underset{2.20}{\implies} Z(A_L) = L\).
            \item If \(A\) is simple and \(Z(A) = K\) then \(A_L\) is simple.
            \item \(A\in \operatorname{CSA}(K) \implies A_L \in \operatorname{CSA}(L)\). 
        \end{enumerate} 
    \end{proposition}

    Note that:

    \[
        \left( \bigoplus_{i\in I} V_i \right) \otimes_K W \cong \bigoplus_{i\in I} \left( V_i \otimes_K W \right) \;.
    \]

    \[
        K \otimes_K V \xrightarrow{\cong} V, \quad \lambda \otimes v \mapsto \lambda v \;.
    \]

    \begin{proof}
        \begin{enumerate}[label=\roman*)]
            \item \(A \underset{K\text{-v.s.}}{\cong} K^{(I)} \implies A \otimes_K L \cong K^{(I)} \otimes_K L \cong (K \otimes_K L)^{(I)}\cong L^{(I)}\).
            
            \((e_i)_{i\in I}\) basis of \(A\) as \(K\)-v.s. \(\implies (a_i \otimes 1)_i\) is a basis of \(A \otimes_K L\) as \(L\)-v.s.

            \item[iii)] \(2.21 \implies A \otimes_K L\) is simple.
            
            \item[iv)] \(\impliedby\) i, ii, iii.
        \end{enumerate} 
    \end{proof}

    \begin{corollary}
        \(A \in \operatorname{CSA}(K) \implies [A:K]\) is a square.
    \end{corollary}

    \begin{proof}
        Let \(C / K\) be an algebraically closed field. Then \(A \otimes_K C \in \operatorname{CSA}(C) \implies A \otimes_K C \underset{HW4}{\cong} M_n(C) \implies n^2 = [A_C : C] = [A:K]\).
    \end{proof}

    \begin{corollary}
        [Definition of Reduced Degree]
        If \(D\) is a \(K\)-division algebra with \(Z(D) = K\), then \([D:K]\) is infinite or a square. In the latter case we set \(s(D) = \sqrt{[D:K]}\) and call it the Schur index of \(D\).

        If \(A \in \operatorname{CSA}(K)\) and \(A \sim_K D\) then \(s(A) \coloneqq s(D)\). 
    \end{corollary}

    \begin{remark*}
        If \(A\in \operatorname{CSA}(K)\) then by Wedderburn \(A \cong_K M_r(D)\).

        \(\implies [A:K] = r^2 \cdot [D:K] = r^2 s^2, s = S(D) = s(A)\).

        \(n \coloneqq \sqrt{[A:K]} = r \cdot s\) is called the \underline{reduced} degree of \(A\). 
    \end{remark*}

    \begin{theorem}
        The binary operation \([A] \cdot [B] = [A \otimes_K B]\) on \(\operatorname{Br}(K)\) is well defined and gives \(\operatorname{Br}(K)\) the structure of an abelian group with identity element \([K]\) and the inverse of \([A]\) is \([A^o]\).
    \end{theorem}

    \begin{proof}
        There are canonical isomorphism of \(K\)-algebras:

        \[
            A \otimes_K B \xrightarrow{\cong} B \otimes_K A ; \quad a \otimes b \mapsto b \otimes a \;.
        \]

        and

        \[
            (A \otimes_K B) \otimes_K C \xrightarrow{\cong} A \otimes (B \otimes C) ; \quad (a \otimes b) \otimes c \mapsto a \otimes (b \otimes c)\;.
        \]

        We note: \(M_n(E) \underset{K\text{-algs}}{\cong} M_n(K) \otimes_K E\) for any \(E \in K\)-Alg:
        
        \[
            (a_{ij}) \mapsfrom \sum e_{ij} \otimes a_{ij}
        \]

        where \(e_{ij}\) is the elementary matrix: \(e_{ij} = (\delta_{ik}\delta_{jl})_{1 \leq k,l \leq n}\).

        If \(A \cong M_n(D_1), B\cong M_m(D_2), D_1 \otimes_K D_2 \cong M_\ell(D_3)\) then,

        \(A \otimes B \cong (M_n(K) \otimes D_1) \otimes (M_m(K) \otimes D_2)\)
        
        \(\cong (M_n(K) \otimes M_m(K)) \otimes (D_1 \otimes D_2)\)

        \(\cong M_n(M_m(K)) \otimes M_\ell(D_3) \cong M_{nm}(K) \otimes M_{\ell}(K) \otimes D_3 \cong M_{mn\ell}(D_3)\).

        Then \([D_1] \cdot [D_2] = [A \otimes B] = [D_3]\).

        2.25 yields the inverse element.
    \end{proof}

    \begin{remark*}

        \begin{enumerate}[label=\arabic*)]
            \item (division algebra) \(\otimes\) (division algebra) in general is not a division algebra.

            For example, \(D \otimes_K D^o \cong M_{[D : K]}(K)\) which is a division algebra if and only if \([D:K] = 1\), which happens when \(D = K\) a field.

            \item \(\mathbb{H} \otimes_\mathbb{R} \mathbb{H} \cong \mathbb{H} \otimes_\mathbb{R} \mathbb{H}^o \cong M_4(\mathbb{R})\) by \(a + bi + cj + dk \mapsto a - bi - cj - dk\).
            \item \(K\) algebraically closed implies \(\operatorname{Br}(K)\) is the trivial group.
            \item \(\vert K \vert < \infty \implies \operatorname{Br}(K)\) is trivial.
        \end{enumerate}  
    \end{remark*}

    \begin{proposition}

        Let \(L / K\) be a field extension. Then the map:

        \[
            \operatorname{res}_{L / K} : \operatorname{Br}(K) \to \operatorname{Br}(L)
        \]

        \[
            [A] \mapsto [A_L] (=A \otimes_K L)
        \]

        called the \underline{restriction map} is a group homomorphism.

        If \(K \subset F \subset L\) then \(\operatorname{res}_{L / K} = \operatorname{res}_{L / F} \circ \operatorname{res}_{F / K}\).

        We define \(\operatorname{Br}(L / K) \coloneqq \ker (\operatorname{res}_{L / K}) = \left\{ [A] \in \operatorname{Br}(K) : [A_L] = [L] \right\}\).
        
        When \([A] \in \operatorname{Br}(L / K)\) then we say that \(L\) is a \underline{splitting field} of \(A\), or that \(A\) \underline{splits over} \(L\).

        \[
            A \text{ splits over } L \iff A_L\cong M_n(L) \text{ as } L \text{-algebras.}  
        \]
    \end{proposition}

    \begin{proof}
        \(\operatorname{res}_{L / K}\) is a homomorphism:

        \[
            (A \otimes_K B) \otimes_K L= A \otimes_K (B \otimes_K L) \cong A \otimes_K (L \otimes_L (B \otimes_K L)) = (A \otimes_K L) \otimes_L (B \otimes_K L)\;.
        \]
    \end{proof}

    \begin{convention*}
        In the beginning of a statement, `\(L / K\)' stands for `Let \(L\) be a field extension of \(K\)'.
    \end{convention*}

    \begin{proposition}
        \(L / K, A \in \operatorname{CSA}(K)\).
        
        \begin{enumerate}[label=\roman*)]
            \item \(s(A_L) \mid s(A)\). 
            \item \(s(A_L) \iff [A] \in \operatorname{Br}(L / K)\).
        \end{enumerate} 
    \end{proposition}

    \begin{proof}
        \(A\cong M_n(D), s(A) = s(D)\).

        \(A_L \cong M_n(D_L), s(A_L) = s(D_L), D_L \cong M_m(D^{\prime})\).

        \(\implies [D:K] = [D_L:L] = m^2[D^{\prime} : L]\)

        \(\implies s(A)=s(D)=\sqrt{[D:K]} = ms(D^{\prime}) = ms(D_L) = ms(A_L)\).
    \end{proof}

    \begin{theorem}
        [Centralizer Theorem]

        Let \(A \in \operatorname{CSA}(K), B \subset A\) a simple \(K\)-subalgebra. Set \( C = Z_A(B)\).

        \begin{enumerate}[label=\roman*)]
            \item \(C\) is a simple \(K\)-subalgebra of \(A\).
            \item \(C\) is similar (as \(K\)-algebra) to \(B^{o} \otimes_K A\): \(C \sim_K B^o \otimes_K A\). Furthermore, \(Z(C) \cong Z(B^o \otimes_K A) \cong Z(B)\).
            \item \(Z_A(Z_A(B)) = Z_A(C) = B\) [main assertion].
            \item \([A:K] = [B:K][C:K]\).
            \item If \(L = Z(B)\) then \(B \otimes_L C \sim_L A_L = A \otimes_K L\) as \(L\)-algebras.
            \item If \(Z(B) = K\), then the map \(B \otimes_K C \to A\), \(b \otimes c \mapsto bc\) is an isomorphism of \(K\)-algebras.  
        \end{enumerate} 
    \end{theorem}

    \begin{remark*}
        2.32 is more generally true if we only assume:

        \begin{itemize}
            \item \(A\) is a simple artinian \(K\)-algebra with center \(K\).
            \item \([B:K] < \infty\). 
        \end{itemize} 
    \end{remark*}

    \section*{Tuesday, 2/17/2026}
    
    \begin{proof}
        2.19 \(\implies \operatorname{End}_{B \otimes_K A^o}(A) = Z_A(B) = C\).
        
        Apply 2.12(ii): \(\mathscr{A} = B \otimes_K A^o\), which is simple by 2.21 and \(\mathscr{M} = A\) as an \(\mathscr{A}\)-module. Then \(\mathscr{M}\) is finitely generated [by \(1\)].

        Then, \(\mathscr{B} = \operatorname{End}_{\mathscr{A}} (\mathscr{M}) = C\) is simple.

        \(C\) is a subalgebra of \(A\) which is finite dimensional over \(K\). Thus, \(C\) is finite dimensional over \(K\). This proves (i).

        Moreover, 2.12(ii) tells us that \(\mathscr{B}\) is similar, as \(K\)-algebra, to \(\mathscr{A}^o\).

        i.e. \(C = \mathscr{B} \sim_K \mathscr{A}^o = (B \otimes A)^o = B^o \otimes_K A\) and \(Z(C) = Z(B^o \otimes_K A) = Z(B^o) \otimes_K Z(A) = Z(B)\), hence (ii). Furthermore, \([A:K]^2 \underset{2.12}{=} [\mathscr{A} : K] [\mathscr{B} : K] = [B:K][A:K][C:K] \implies [A:K] = [B:K][C:K]\), hence (iv). 

        Set \(L = Z(B) \overset{ii}{=} Z(C) \implies B \otimes_L C \overset{\text{check!!!}}{\sim_L} B \otimes_L (B^o \otimes_K A) \cong (B \otimes_L B^o) \otimes_K A \sim L \otimes_K A = A_L\), hence (v).
        
        Set \(B^{\prime} = Z_A(C)\). Elements in \(B\) commute with \(C\),thus \(B \subset Z_A(C)\). \(C\) is simple (by (i)). Then apply part i on the theorem where we set \(A\) to be \(A\) and \(B\) to be \(C\). Then \(B^{\prime}\) must also be simple. iv \(\implies [B:K][C:K] = [A:K] = [C:K][B^{\prime}:K] \implies [B:K]=[B^{\prime}:K]\). 

        This coupled with the fact that \(B \subset B^{\prime}\) implies \(B = B^{\prime}\), hence (iii).

        Assume \(Z(B) = K\) and consider \(B \otimes_K C \xrightarrow{\gamma} A, a \otimes b \mapsto ab\). Note that \(2.21 \implies B \otimes_K C\) is simple, thus \(\ker (\gamma) = 0 \implies \gamma\) is injective. iv \(\implies \gamma\) is an isomorphism of \(K\)-algebras, hence (vi). 
    \end{proof}

    \begin{theorem}
        Let \(A \in \operatorname{CSA}(K)\) and \(L \subset A\) a \(K\)-subalgebra which is a field. TFAE:

        \begin{enumerate}[label=\roman*)]
            \item \(L\) is a maximal commutative subalgebra, i.e. \(Z_A(L) = L\). 
            \item  \([A:K] = [L:K]^2\).
        \end{enumerate} 

        If (i) or (ii) is satisfied, then \(A_L \sim_L L\), i.e. \(L\) is a splitting field of \(A\).
    \end{theorem}

    \begin{proof}
        Apply the Centralizer Theorem (2.32) with \(B = L\) and \(C = Z_A(L)\).

        iv \(\implies\) if \(L = Z_A(L)\) then we have \([A:K]=[B:K][C:K]=[L:K]^2\quad(\ast)\).
        
        Conversely, if \([B:K][C:K] = [L:K][C:K] = [A:K] = [L:K]^2 \implies [L:K][C:K]\). Hence, the inclusion \(L \subseteq C \implies L = C\). 
        
        By 2.32(v), \(L \otimes_L L = L \sim A \otimes_K L = A_L\).
    \end{proof}

    \begin{remark*}
        Not every \(A \in \operatorname{CSA}(K)\) has a \(K\)-subalgebra \(L\) which is a field and satisfies 2.33(i) or (ii).

        E.g. suppose \(K\) is algebraically closed and \(A = M_n(K)\) with \(n > 1\).
    \end{remark*}

    \begin{theorem}
        Let \(D \in \operatorname{CSA}(K)\) which is a division algebra. Then \(D\) has maximal subfields, and any of them is a splitting field of \(D\).

        For every maximal subfield \(L\) of \(D\), we have that \(K \subset L\), and \([L : K] = s(D)\).
    \end{theorem}

    \begin{proof}
        A maximal subfield \(L\) of \(D\) is commutative, and contains \(K\) [consider the composite field \(K . L \supset L\).] Such maximal subfields exist since \([D:K] < \infty\).

        Given such an \(L\) we have \(L \subseteq Z_D(L) \eqqcolon C\). For \(c\in C\), \(L[c] =\) smallest \(K\)-subalgebra of \(D\) containing \(L\) and \(c\) must be commmutative [since \(c\) commutes with alll elements of \(L\)], is finite dimensional, and still a subfield of \(D\).

        [Any non-zero element \(d \in L[c]\) acts injectively by multiplication on \(L[c]\) since \(D\) is a division algebra. \([D:K] < \infty \implies [L[c] : K] < \infty \implies\) multiplication by \(d\) is surjective, hence \(d\) is invertible in \(L[c]\).]

        However, maximality of \(L\) implies that \(L[c] = L\). Therefore, \(Z_D(L) = L\).

        2.33 implies that \(L\) is a splitting field of \(D\).

        2.33 also implies that \([L:K] = \sqrt{[D:K]} = s(D)\). 

    \end{proof}

    \begin{theorem}
        \(\forall A \in \operatorname{CSA}(K), \exists\) splitting field \(L\) such that \([L:K] = s(A)\).
    \end{theorem}

    \begin{proof}
        \(A \sim_K M_n(D)\). Now use 2.34 and \(s(A) = s(D)\), by definition.
    \end{proof}

    \begin{theorem}
        Let \(A \in \operatorname{CSA}(K)\). If \(L / K\) is a splitting of \(A\) with \([L : K] < \infty\), then \(\exists A^{\prime} \in \operatorname{CSA}(K)\) such that \(A^{\prime} \sim_K A\) and \(L\) is a maximal commutative subalgebra of \(A^{\prime}\).
    \end{theorem}

    \begin{proof}
        \(A_L \cong_L M_n(L) = \operatorname{End}_L(V), V = L^n\).

        Note: \(A \hookrightarrow A \otimes_K L, a \mapsto a \otimes 1\). But \(A \otimes_K L \simeq \operatorname{End}_L(V) \hookrightarrow \operatorname{End}_K(V) \eqqcolon \mathscr{A} \in \operatorname{CSA}(K)\). \(\mathscr{A} \cong M_{nd}(K)\) where \(d = [L:K]\). \([V:K] = n[L:K]\).

        Set \(\mathscr{B} \coloneqq A, \mathscr{C} \coloneqq Z_{\mathscr{A}} (\mathscr{B})\). Apply 2.32(ii).

        \(\mathscr{C} \sim_K \mathscr{B}^o \otimes_K \mathscr{A} = A^o \otimes_K \operatorname{End}_K(V) \sim A^o\).

        2.32(iv) implies that \([\mathscr{C} : K] = \frac{[\mathscr{A} : K]}{[\mathscr{B} : K]} = \frac{[V:K]^2}{[A:K]} = \frac{n^2 [L:K]^2}{[A_L : L]} = [L:K]^2\). 

        Note: \(L \hookrightarrow \mathscr{A}, \lambda \mapsto [v \mapsto \lambda v]\) and elements in \(L\) commute with elements in \(\operatorname{End}_L(V)\), hence with elements in \(A\), this gives \(L \hookrightarrow Z_{\mathscr{A}}(\mathscr{B}) = \mathscr{C}\).

        By 2.33, \(L\) is a splitting field of \(\mathscr{C}\).

        \(\implies L\) is also a splitting field of \(\mathscr{C}^o \sim_K A\).

        Take \(A^{\prime} \coloneqq \mathscr{C}^o\).
    \end{proof}

    \begin{theorem}
        Let \(A \in \operatorname{CSA} (K)\) have Schur index \(s\).

        If \(L / K\) is a splitting field of \(A\), then \(s \mid [L:K]\).

        \begin{enumerate}[label=\roman*)]
            \item If \(L / K\) is a splitting field of \(A\), then \(s \mid [L:K]\).
            \item \(A\) has splitting fields \(L\) where \([L:K] = s\).
        \end{enumerate} 

         The splitting fields of \(A\) of degree \(s\) over \(K\) are precisely those \(L / K\) which are \(K\)-isomorphic to a maximal subfield of a division algebra \(D \sim_K A\).
        
    \end{theorem}

    \begin{proof}
        Let \(L / K\) be a splitting field of \(A\) with \([L:K] < \infty\). Assume that \(A \sim_K D\).

        2.36 \(\implies \exists r \geq 1\) such that \(L\) is a maximal commutative subalgebra of \(M_r(D)\).

        2.33 \(\implies [L:K]^2 = r^2[D:K] \implies \frac{[L:K]}{s(D)} = r\), hence (i).

        If \([L:K] = s\), \(r = 1\) and \(L \subset D\) is a maximal subfield.

        Conversely, if \(L\) is isomorphic to a maximal subfield of \(D \sim_K A\), then by 2.34, \(L\) is a splitting field of \(D \implies L\) is a splitting field of \(A\), and \([L:K] \underset{2.34}{=} s(D) = s(A)\).
    \end{proof}

    \section*{Thursday, 2/19/2026}
    
    \begin{remark*}
        In the proof of the centralizer theorem, we used (ii) \(C = Z_A(B) \sim_K B \otimes_K A^o.\)

        In the proof of (v), we set \(L = Z(B) \cong Z(C)\) and used that \(L\) is simillar to \(B \otimes_K A^o\) as \(L\)-algebras.

        This is indeed the case.

    \end{remark*}

    \begin{proposition}
        \(A \in \operatorname{CSA}(K), L / K\) fiinite extension.

        \begin{enumerate}[label=\roman*)]
            \item \(s(A) \mid [L:K]s(A_L)\).
            \item If \(\operatorname{gcd}([L:K] , s(A)) = 1\), then \(s(A) = s(A_L)\).  
        \end{enumerate} 
    \end{proposition}

    \begin{proof}
        \begin{enumerate}[label=\roman*)]
            \item Let \(E / L\) be a splitting field of \(A_L\) with \([E:L] = A_L\). \(E\) exists by 2.35.
            
            For some \(n\), \(M_n(E) \cong A_L \otimes_L E = (A \otimes_K L) \otimes_L E = A \otimes_K E \implies E\) is a splitting field of \(A\).

            2.37 \(\implies s(A) \mid [E:K] = [E:L][E:K] = [L:K] s(A_L)\).

            \item \(s(A) \mid s(A_L)\) by i. By 2.31: \(s(A_L) \mid s(A)\). Therefore, \(s(A) = s(A_L)\).
        \end{enumerate} 
    \end{proof}

    \begin{lemma}
        Let \(D \in \operatorname{CSA}(K)\) be a division algebra. If \(s(D) > 1\) [i.e. \(D\) is not \(K\) itself], then \(D\) contains subfields \(E / K\) which are proper separable extensions of \(K\).
    \end{lemma}

    \begin{proof}
        HW6.
    \end{proof}

    \begin{proposition}
        Let \(D\in \operatorname{CSA}(K)\) be a division algebra. Among the maximal subfields of \(D\), there are some \(L\) which are separable over \(K\). Hence every \(A\in \operatorname{CSA}(K)\) has a splitting field \(L\) such that:

        \begin{enumerate}[label=\roman*)]
            \item \(L / K\) is separable.
            \item \([L:K] = s(A)\). 
        \end{enumerate} 
    \end{proposition}

    \begin{proof}
        Fields of characteristic \(0\) only have separable extensions, thus the statement is trivially true.
 
        Suppose \(\operatorname{char} (K) = p > 0\). Let \(L \subset D\) be a subfield which is maximal among all separable extensions of \(K\).

        WTS: \(L = Z_D(L)\). Then \(L\) is maximal among \underline{all} subfields of \(D\).
        
        Set \(C\coloneqq Z_D(L)\). Apply Centralizer theorem with the setting:

        \[
            A = D \quad B = L \quad C = C\;.
        \]

        Then \(Z(B) = Z(C) = Z(L) = L\).

        Furthermore, \(C\) is simple \(\implies C \in \operatorname{CSA}(L)\) and \(C\) is a division algebra.

        Claim: \(C\) cannot contain a separable extension \(E \supsetneq L\).

        Such an \(E\) would be a separable extension of \(K\) which is strictly larger than \(L\). This contradicts the maximality of \(L\).

        Now apply the lemma. 2.39 \(\implies C = L\). Hence \(L = Z_D(L)\) and thus a maximal subfield of \(D\).

    \end{proof}

    \begin{corollary}
        Every \(A \in \operatorname{CSA}(K)\) has a splitting field which is a finite Galois exxtension. In particular,

        \[
            \operatorname{Br}(K) = \bigcup_{L / K \text{ finite Galois}} \operatorname{Br}(L / K).
        \]
    \end{corollary}

    \begin{proof}
        2.40 \(\implies \exists\) splitting field \(L_0 / K\) finite and separable. Let \(L\) be the normal closure of \(L_0\), which is finite, separable and hence Galois \(/ K\).
    \end{proof}

    Question: Does every division algebra \(D \in \operatorname{CSA}(K)\) have a maximal subfield \(L\) which is Galois over \(K\)?

    Answer: No, in general (Amitsur, 1972). But certain fields (e.g. all locally compact fields) have the property that the answer is `yes'.

    Goal: Understand \(\operatorname{Br}(L / K)\) for \(L / K\) \underline{finite Galois}.

    \begin{theorem}
        [Skolem-Noether]

        Let \(A, B \in K\text{-Alg}\) be simple, and assume

        \begin{itemize}
            \item \(A\) is artinian and \(Z(A) = K\).
            \item \([B:K] < \infty\). 
        \end{itemize} 

        Then, for any two \(K\)-algebra homomorphisms \(f,g : B \to A\), there exists \(u\in A^\times\) such that \(x\forall \in B: g(x) = u ^{-1} f(x) u\).

    \end{theorem}

    \begin{proof}
        Consider \(\mathscr{A} = A \otimes_K B^o\) which is simple by 2.21, and artinian by 2.22.

        Define on \(A\) two \(\mathscr{A}\)-module structures:

        \[
            A_f : (\underset{\in A}{a} \otimes \underset{\in B^o}{b}) \underset{f}{\cdot} \underset{\in A}{x} = ax f(b)\;. 
        \]

        \[
            A_g : (\underset{\in A}{a} \otimes \underset{\in B^o}{b}) \underset{g}{\cdot} x = a x g(b)\;.
        \]

        Note: \(A_f\) and \(A_g\) are generated by \(1_A\) as \(\mathscr{A}\)-modules.

        Let Let \(\mathscr{N}\) be a simple \(\mathscr{A}\)-module.

        By 2.3: \(\mathscr{N}\) is up to isomorphism the only simple \(\mathscr{A}\)-module.

        Hence \(A_f \cong_{\mathscr{A}} \mathscr{N}^m , A_g \cong_{\mathscr{A}} \mathscr{N}^n\).

        Note: \(A \to \mathscr{A} , a \mapsto a \otimes 1\) is a \(K\)-Alg homomorphism.

        \(A_f\) is finitely generated by \(1_A\) as an \(A\)-module. By projecting \(\mathscr{N}^m\) onto the first summand, we get \(\mathscr{N}\) which is finitely generated as an \(A\)-module.

        \(A\) is semisimple and artinian, it is semisimple. Let \(N\) be a simple \(A\)-module. Then, \(\mathscr{N} \cong N^r\) as an \(A\)-module.

        Therefore, \(rm = \ell_A(\underset{\cong N^{\oplus rm}}{\mathscr{N}^m}) = \ell_A(A_f) = \ell_A(A_g) = \ell_A(\mathscr{N}^n) = rn\).
        
        Therefore, \(m = n\).

        Thus \(\exists \mathscr{A}\)-module isomorphism \(\varphi : A_f \xrightarrow{\cong} A_g\). Set \(u \coloneqq \varphi(1)\).

        \[
            ug(\underset{\in B}{x}) = \varphi (1) g(x) = (1 \otimes x) \underset{g}{\cdot} \varphi(1) = \varphi((1 \otimes x) \underset{f}{\cdot} 1) = \varphi (f(x))
        \]
        \[    
            = \varphi(f(x) \cdot 1) = \varphi((f(x) \otimes 1) \underset{f}{\cdot} 1) = (f(x) \otimes 1) \underset{g}{\cdot} \varphi (1) = f(x) \varphi (1) \cdot 1 = f(x) u 
        \]

        Multiplying by \(u ^{-1} \) in the right, we see that \(u g(x) u ^{-1} = f(x)\), which is what we wanted.

        Exercise: \(u \in A^\times\).

    \end{proof}

    \begin{definition}
        Let \(A\) be an algebra. An \underline{inner automorphism} of \(A\) is an automorphism of the form \(x \mapsto u ^{-1} x u\) for some \(u \in A^\times\).
    \end{definition}

    \begin{theorem}
        \(A \in \operatorname{CSA}(K), g \in \operatorname{Aut}_{K\text{-Alg}}(A) \implies g \) is inner.

        Hence \(\operatorname{Aut}_{K \text{-Alg}}(A) \cong A^\times / K^\times\). 
    \end{theorem}

    \begin{proof}
        Apply 2.42 with \(B=A , f = \operatorname{id}_{A}, g = g \implies \exists u\in A^\times : \forall x\in A: g(x) = u ^{-1} x u\).

        Thus \(A^\times \to \operatorname{Aut}_{K\text{-Alg}}(A), u \mapsto [x \mapsto u x u ^{-1}]\) is surjective, and its kernel is \(K^\times\). 

    \end{proof}

    \begin{theorem}
        Let \(A\in K\text{-Alg}\) be simple and artinian, \(Z(A) = K, B \subset A\) simple subalgebra, \([B:K] < \infty\).
        
        Let \(N = \{ u \in A^\times : u B u ^{-1} = B \}, C = Z_A(B)\).

        Then the map \(N / C^\times \to \operatorname{Aut}_{K\text{-Alg}}(B), u C^\times \mapsto [x \mapsto u x u ^{-1}]\) is an isomorphism.
    \end{theorem}

    \begin{theorem}
        [Wedderburn] Every finite division algebra is a finite field. Hence, \(\operatorname{Br}(K)\) is trivial if \(\vert K \vert < \infty\). 
    \end{theorem}

    \begin{proof}
        [Sketch]

        Let \(K = \mathbb{F}_q , D \in \operatorname{CSA}(K)\) a division algebra.

        Let \(\mathscr{L} = \left\{ L \subset D : \text{maximal subfields} \right\} \).

        Theorem 2.34 \(\implies \forall L \in \mathscr{L} : [L:K] = s(D) = s\).

        Then, \(\forall L \in \mathscr{L}, L \cong \mathbb{F}_{q^s}\).

        Fix \(L\in \mathscr{L}\).

        Skolem-Noether \(\implies \forall L_1 \in \mathscr{L}, \exists u\in D^\times : u ^{-1} L u = L_1\).
        
        Note: \(\forall x \in D, \exists L_x \in \mathscr{L} : x \in L_x \implies D^\times = \bigcup_{u\in D^\times} u ^{-1} L^\times u\) 

        Note: \(D^\times \twoheadrightarrow \mathscr{L}, u \mapsto u ^{-1} L u\).

        \(N \coloneqq \operatorname{Stab}_{D^\times}(L) = \left\{ u \in D^\times : u ^{-1} L u \equiv  L \right\} \supset L^\times\).

        \(m \coloneqq \vert \mathscr{L} \vert = \vert D^\times  \vert / \vert N \vert \leq \vert D^\times  \vert / \vert L^\times \vert\).

        \(m \vert L^\times \vert = \vert \mathscr{L} \vert \vert L^\times \vert  = \bigsqcup_{L_1 \in \mathscr{L}} \vert L_1^\times \vert \underset{\text{if } m > 1}{>} \left\vert \bigcup_{L_1 \in \mathscr{L}} L_1^\times \right\vert = \vert D^\times \vert \)
        
        Therefore, \(m > \frac{\vert D^\times \vert}{\vert L^\times \vert}\).

        This is a contradiction.

    \end{proof}

    \section*{Tuesday, 2/24/2026}
    
    \section*{Thursday, 2/26/2026}
    
    \section*{Tuesday, 3/3/2026}
    
    \section*{Thursday, 3/5/2026}

    \section*{Tuesday, 3/10/2026}
    
    \section{for numbering}

    \begin{theorem}
        for numbering
    \end{theorem}

    \subsection*{Tensor Products of Representations}

    Works for group algebras \(KG\) since these are Hopf algebras.

    Let \((R,V)\) and \((S,W)\) be representations of \(G\).

    Define \(R \otimes S \coloneqq T : G \to GL(V \otimes_K W)\) by:

    \[
        T(g)(v \otimes w) = R(g)v \otimes S(g) w\;.
    \]

    \[
        \left( \sum a_g g \right) (v \otimes w) = \sum a_g \left( R(g) \cdot v \otimes S(g) \cdot w \right) \;.
    \]

    Reason why this works: \(KG \to KG \otimes_K KG, \sum a_g g \mapsto \sum a_g(g \otimes g)\) is a \(K\)-algebra homomorphism.

    Examples:

    \begin{enumerate}[label=\arabic*)]
        \item Let \(G = \mathbb{Z} / n\mathbb{Z}\) and \(\chi_{\lambda_1} , \chi_{\lambda_2} : G \to \mathbb{C}^\times = GL(\mathbb{C})\), \(\overline{a} \mapsto \lambda_i^a\), \(\lambda_i^n = 1\).
        
        Then \(\chi_{\lambda_1} \otimes \chi_{\lambda_2} : \overline{a} \mapsto (\lambda_1 \lambda_2)^a\).

        i.e. \(\chi_{\lambda_1} \otimes \chi_{\lambda_2} = \chi_{\lambda_1 \lambda_2}\).

        \item \(K^3 = Ke_1 \oplus Ke_2 \oplus Ke_3\). Let \(L = K(e_1 + e_2 + e_3)\) and consider \(K^3 / L\).
        
        \(\mathscr{S}_3\) has a unique irreducible \(2\)-dimensional representation \(V\) over any field \(K\) of \(\operatorname{char}(K) \nmid 6 \implies \operatorname{sgn} \otimes V \cong V\).
    \end{enumerate} 

    \begin{lemma}
        If \((R,V)\) and \((S,W)\) are finite dimensional \(G\)-reps, then \(\chi_{V \otimes W} = \chi_V \cdot \chi_W\).
    \end{lemma}

    \begin{proof}
        Choose basis \(v_1 , \cdots , v_n\) of \(V\) and \(w_1 , \cdots ,w_m\) of \(W\).
        
        Then, \(R(g) \leftrightarrow \begin{bmatrix}
            a_{11} (g) & \cdots & a_{n1} (g) \\
            \vdots &  & \vdots \\
            a_{1n} (g) & \cdots  & a_{nn} g \\
        \end{bmatrix}, S(g) \cong \text{matrix} \in M_m(K)\).

        Then action of \(R(g) \otimes S(g)\) on the basis \(v_1 \otimes w_1 , v_1 \otimes w_2 , \cdots , v_1 \otimes w_m , v_2 \otimes w_1, \cdots , v_2 \otimes w_m, \cdots , v_n \otimes w_1, \cdots , v_n \otimes w_m\) is given by:

        \[
            \begin{bmatrix}
                \boxed{a_{11}(g) S(g)} & \cdots & \boxed{a_{1n}(g) S(g)} \\
                \boxed{a_{21}(g) S(g)} & \cdots & \vdots \\
                \vdots &  & \vdots \\
                \boxed{a_{n1}(g) S(g)} & \cdots & \boxed{a_{nn}(g) S(g)} \\
            \end{bmatrix} \;.
        \]

        Evidently, each box is \(m \times m\).

        Then, the trace of this matrix is given by:

        \[
            a_{11} (g) \Tr(S(g)) + a_{22}(g) \Tr(S(g)) + \cdots + a_{nn}(g) \Tr (S(g)) = \chi_V(g) \cdot \chi_W(g)\;.
        \]
    \end{proof}

    Example: If \(V, \operatorname{sgn}\) are the representations of \(\mathscr{S}_3\) in the example, then:

    \[
        \chi_V = \chi_{\operatorname{sgn} \otimes V} \overset{3.2}{=} \operatorname{sgn} \cdot \chi_V\;.
    \]

    Therefore, \(\forall\) transposition \(\tau\in \left\{ (12),(23),(13)\right\}\), \(\chi_V(\tau) = - \chi_V(\tau)\).
    
    We have assumed that the characteristic is not \(2\). Therefore, \(\chi_V(\tau) = 0\).

    \begin{proposition}
        Let \(G\) be a finite group of order \(n\).

        For the character \(\rho = \chi_{KG}\) of \(G\) on \(KG\) (by left multiplication), we have:

        \[
            \rho(g) = \begin{dcases}
                n \cdot 1_K, &\text{ if } g = 1_G ;\\
                0, &\text{ otherwise} .
            \end{dcases}
        \]
    \end{proposition}    

    \begin{proof}
        We have already seen:

        \[
            \Tr(1_G \mid V) = [V:K] \cdot 1_K\;.
        \]

        Write \(G = \{ g_1 , \cdots , g_n \}\). Then \(g_1 , \cdots , g_n\) is a \(K\)-basis of \(KG\). The multiplication-by-\(g\) map:

        \[
            \sum_{i=1}^n a_i g_i \mapsto \sum_{i=1}^n a_i \cdot (g g_i)\;.
        \]

        is given by a permutation matrix, the diagonal entries of which are only zeroes if \(g \neq 1_G\).

        Indeed, the map is \(g_i \mapsto g g_i = g_{\pi(i)}\), where \(\pi\) is a non-trivial permutation. We can have a non-zero diagonal entry if and only if \(g g_i = g_i \implies g = 1_G\).
        
        Therefore, \(\chi_{KG}(g) = 0\).

    \end{proof}

    \begin{definition}
        A function \(f: G \to K\) is called a \underline{class function} if it is constant on every conjugacy class of \(G\).

        i.e. \(\forall g,h \in G: f(hgh ^{-1}) = f(g)\). We set:

        \[
            Z_K G = \left\{ f: G \to K \mid f \text{ is a class function}  \right\} \left( \subseteq K^G \right)\;.
        \]

        This is a \(K\)-algebra with pointwise addition and multiplication, and scalar multiplication: \((f_1 + f_2)(g) = f_1(g) + f_2(g), (f_1 \cdot f_2)(g) = f_1(g) \cdot f_2(g), (c \cdot f)(g) = cf(g)\).

    \end{definition}

    Example: If \(G\) has only finitely many conjugacy classes \(C_1 , \cdots , C_h\) [e.g. \(\vert G \vert < \infty\)], then:

    \[
        Z_K(G) \underset{K\text{-alg}}{\cong} \underbrace{K \times K \times \cdots \times K}_{h \text{ copies}}.
    \]

    Caution: If \(G\) is finite, then \(Z_K(G)\) can be considered as a subspace of \(KG\), but the multiplication is different.


    Example: Let \(V\) be a finite dimensional representation of \(G\). Then, \(\chi_V : G \to K\) is a class function.

    If \(\mathbf{1}\) is the trivial representation on \(K\) [every \(g\in G\) acts by \(1\)], then \(\chi_{\textbf{1}}(g) = 1\) for every \(g\in G\).
    
    \section*{Thursday, 3/12/2026}
    
    \begin{definition}
        The \underline{character ring} \(X_K(G)\) consists of all functions \(f: G \to K\) such that \(\exists\) finite-dimensional / finite-degree representations \(V_1 , \cdots , V_n \in \operatorname{Rep}_K G^{\text{f.d.}}\) and integers \(a_1 , \cdots , a_n \in G\) such that:
        
        \[
            f = a_1 \chi_{V_1} + \cdots + a_n \chi_{V_n} \;.
        \]

        Such functions are trivially class functions: recall that characters are trace functions, and \(\Tr(B A B ^{-1}) = \Tr (A)\).

        Hence \(H_K G \subset Z_K G\), and we equip this abelian group with the induced ring structure (pointwise multiplication). It is indeed a ring, since lemma 3.2 states that \(\chi_V \cdot \chi_W = \chi_{V \otimes W}\).

    \end{definition}

    \begin{remark*}

        Recall that \(\chi_{V_1} + \chi_{V_2} = \chi_{V_1 \oplus V_2}\).

        We can then write such a sum \(f\) as follows:

        \[
            f = \sum_{a_i \geq 0} a_i \chi_{V_i} - \sum_{a_j < 0} (-a_j) \chi_{V_j} = \chi_{\bigoplus_{a_i \geq 0} V_i^{\oplus a_i}} - \chi_{\bigoplus_{a_j < 0} V_j^{\oplus (-a_j)}}\;.
        \]

        These \(f\in X_K G\) are called \underline{virtual characters} [or characters of virtual representation].
    \end{remark*}

    \begin{proposition}
        missed?
    \end{proposition}

    \begin{proposition}
        Any \(f\in X_K G\) is a class function.
    \end{proposition}

    \begin{proof}
        Easy.
    \end{proof}

    \begin{lemma}
        Suppose \(g\in G\) satisfies \(g^m\) for some \(m > 0\), and let \(V\in \operatorname{Rep}_K G^{\text{f.d.}}\).

        Then \(\exists\) roots of unity \(\xi_1 , \cdots , \xi_d \in \overline{K}\) [which is the algebraic closure of \(K\)] such that \(\xi_1^d = \cdots =\xi_d^m = 1\) and \(\chi_V(g) = \xi_1 + \cdots + \xi_d\). 
    \end{lemma}

    \begin{proof}
        Suppose the representation \(V\) is actually \((T,V)\). \(g^m = 1 \implies T(g)^m =\operatorname{id}_{V}\). Hence the minimal polynomial of \(T(g)\) divides \(t^m - 1\). Then, all eigenvalues \(\xi\) of \(T(g)\) in \(\overline{K}\) [which are the roots of the characteristic polynomial] are the roots of the minimal polynomial, hence they satisfy \(\xi^m = 1\).
        
        If \(\xi_1 , \cdots , \xi_d\) are the eigenvalues of \(T(g)\) in \(\overline{K}\) (with multiplicity), then, 
        
        \[
            \chi_V(g) = \Tr(T(g)) = \xi_1 + \cdots + \xi_d\;.
        \]
    \end{proof}

    \begin{definition}
        Let \(T: A \to \operatorname{End}_K (V)\) be a representation of any \(A\in K\)-Alg.

        \begin{enumerate}[label=\arabic*)]
            \item For any field extension \(E / K\) we obtain a representation: \(T^E : A_E \coloneqq A \otimes_K E \to \operatorname{End}_E(V_E)\) by `scalar extension':
            
            \[
                T^E(\underset{\in A}{a} \otimes \underset{\in E}{c})(\underset{\in V}{v} \otimes \underset{\in E}{c^{\prime}}) = T(a) \cdot v \otimes cc^{\prime}\;.
            \]

            \((T^E , V_E = V \otimes_K E)\) is called the \underline{base change} of \((T,V)\) to \(E\).

            Note: When \([V:K] < \infty\), then \(\chi_{T^E}(a \otimes 1) = \chi_T(a)\) for all \(a\in A\).

            Example: \(\operatorname{char} (K) = p, g^p = 1 \implies \chi_V(g) = [V:K] \cdot 1_K\).

            Notation: We set \(\chi_T^E \coloneqq \chi_{T^E}\).

            \item \(T: A \to \operatorname{End}_K(V)\) is called \underline{absolutely irreducible} [abs. irred.] if \((T^E, V_E)\) is irreducible for every field extension \(E / K\).
            
            Character \(\chi_T\) is called \underline{abs. irred.} if \(T\) is absolutely irreducible.

        \end{enumerate} 
    \end{definition}

    Examples:

    \begin{enumerate}[label=\arabic*)]
        \item Every \(1\)-dimensional representation is absolutely irreducible.
        \item Any irreducible representation of \(\mathscr{S}_3\) over any field \(K\) of \(\operatorname{char}(K) \neq 2,3\) is absolutely irreducible [HW8: list of irreducible representations is \(\mathbf{1}, \operatorname{sgn},\) unique (up to isomorphism) \(2\)-dimensional representations. These stay irreducible after base change].
        
        Analogous statement is true for \(\mathscr{S}_n\) in general, condition is \(\operatorname{char}(K) \nmid n!\).
        \item \(G = \mathbb{Z} / n\mathbb{Z}, n \geq 3, \rho : G \to \operatorname{GL}_2(\mathbb{R}), \overline{a} \mapsto \begin{bmatrix}
            \cos \frac{2 \pi a}{n} & - \sin \frac{2 \pi a}{n} \\
            \sin \frac{2 \pi a}{n} & \cos \frac{2\pi a}{n}\\
        \end{bmatrix} = T_1(\overline{a})\) is not absolutely irreducible: \(\rho^\mathbb{C} \cong \chi_{\xi_n} \oplus \chi_{\xi_n ^{-1}}, \chi_n = e^{\frac{2\pi i}{n}}\).
    \end{enumerate} 

    \begin{theorem}
        Let \((T,V)\) be a f.d. representation of \(A\) over \(K\).

        \begin{enumerate}[label=\roman*)]
            \item \(T\) is absolutely irreducible \(\iff T(A) = \operatorname{End}_K(V)\).
            \item \(T\) is absolutely irreducible \(\implies \operatorname{End}_A(V) = K \cdot \operatorname{id}_{V}\).
            \item \(V\) is semisimple as an \(A\)-module and \(\operatorname{End}_A(V) = K \cdot \operatorname{id}_{V} \implies T\) is absolutely irreducible.
            \item \(T\) is irreducible and \(K = \overline{K} \implies T\) is absolutely irreducible.
        \end{enumerate} 
    \end{theorem}

    \begin{proof}

        \ \

        \begin{enumerate}[label=\roman*)]
            \item[i\(\implies\):] Suppose \(T(A) \subsetneq \operatorname{End}_K(V) \implies\) for any \(E / K\):
            
            \(T^E(A_E) = \operatorname{span}_E(T(A)) \subsetneq \operatorname{End}_E(V_E)\). Take \(E = \overline{K}\) and set \(W = V_E, B = A_E\).

            \(T\) abs. irred. so \(W\) is a simple \(B\)-module.
            
            Claim: The map \(B \to \operatorname{End}_E(W)\) is surjective.

            Proof: Let \(D \coloneqq \operatorname{End}_B(W)\). Since \(W\) is simple, \(D\) must be a division algebra. \(\operatorname{End}_B(W) \subseteq \operatorname{End}_E(W)\). So, \(D\) is a division algebra which is finite dimensional over \(E\). \(E\) is algebraically closed. Therefore, \(D = E\).

            Jacobson Density Theorem: \(B \to \operatorname{End}_{\operatorname{End}_B(W)}(W) = \operatorname{End}_E(W)\), and since \([W:E] < \infty\), this map is surjective. Thus the claim is proven.

            This claim contradicts that \(T^E(A_E) \subsetneq \operatorname{End}_E(V_E)\). Hence i is proven.

            \item[i \(\impliedby\):] \(T(A) = \operatorname{End}_K(V) \implies V\) is simple as an \(A\)-module.
            
            Choose any \(E / K\). Note that \(T^E(A_E) = \operatorname{End}_E(V_E) \implies V_E\) is simple as an \(A_E\)-module.

            \(\implies T\) is abs. irred.

            \item[ii :] \(T\) abs. irred \(\underset{\text{(i)}}{\implies} T(A)=\operatorname{End}_K(V) \implies \operatorname{End}_A(V) = K \cdot \operatorname{id}_{V}\).
            
            \item[iii :] Assume \((T,V)\) is semisimple. The map in Jacobson's Density Theorem is the map:
            
            \[
                A \to \operatorname{End}_{\operatorname{End}_A(V)}(V) = \operatorname{End}_K(V)\;.
            \]

            which is surjective. i \(\implies T\) is absolutely irreducible.

            \item[iv :] \(T\) irreducible. \(\overset{\text{Schur}}{\implies} \operatorname{End}_A(V) \eqqcolon D\) is a division algebra over \(K = \overline{K} \implies D = K \overset{\text{iii}}{\implies} T\) is abs. irred. 
        \end{enumerate} 
    \end{proof}

    \begin{definition}
        \begin{enumerate}[label=\roman*)]
            \item Given \(A\in K\)-Alg, we call an extension \(E / K\) a splitting field of \(A\) if every f.d. irred. rep's of \(A_E\) is abs. irred.
            
            \begin{remark*}
                When \(A \in \operatorname{CSA}(K)\), then \(E / K\) is a splitting in the sense defined above iff \(E\) is a splitting in the sense of ch. 2 [HW9].
            \end{remark*}

            \item If \(G\) is a group, we call \(E\) a \underline{splitting field of \(G\)}. if all irr. rep's of \(G\) over \(E\) (equiv. if all simple \(EG\)-modules) are abs. irred.
        \end{enumerate} 
    \end{definition}

    Example:

    \begin{enumerate}[label=\arabic*)]
        \item \(G = \mathbb{Z} / n\mathbb{Z}, n \geq 3, K = \mathbb{R}\). A splitting for \(\mathbb{R} G\) is \(\mathbb{C}\). \(\mathbb{R}\) is not a splitting field of \(G\), since \(G\) has irreducible \(2\)-dimensional representations. But the complex irreducible rep's of \(G\) are all \(1\)-dimensional, so \(\mathbb{C}\) is a splitting field.
        \item \(G = \mathbb{Z} / n\mathbb{Z}, n \geq 3, K = \mathbb{Q}\), a splitting field for \(\mathbb{Q} G \cong \mathbb{Q}[T] / (T^n - 1)\) is \(\mathbb{Q}(\xi_n)\) where \(\xi_n\) is a primitive \(n\)-th root of unity. Recall that \(\mathbb{Q}[T]/(T^n - 1) = \prod_{d\mid N} \mathbb{Q}[T] / (\Phi_d(T))\). Note that \(\Phi_d(T) = \prod_{\zeta \text{ prim. \(d\)'th root of unity}} (T- \zeta)\). Hence \(\mathbb{Q}(\zeta_n)\), where \(\zeta_n\) is a primitive \(n\)'th root of \(1\), is a splitting field.  
        \item \(\mathbb{F}_q\) is a splitting field of \(\mathbb{Z} / (q-1)\mathbb{Z}\) [HW10].
        \item Any field \(K\) of char \(\neq 2,3\) is a splitting field of \(\mathscr{S}_3\).  
    \end{enumerate} 

    \section*{Tuesday, 3/24/2026}
    
    \begin{proposition}
        Let \(A\) be any commutative \(K\)-algebra. Then,
        
        \begin{enumerate}[label=\roman*)]
            \item every absolutely irreducible finite-dimensional representation is \(1\)-dimensional.
            
            \item TFAE:
            
            \begin{enumerate}[label=\alph*)]
                \item \(K\) is a splitting field of \(A\).
                \item Every irreducible f.d. represantation of \(A / K\) is one dimensional. 
            \end{enumerate} 

        \end{enumerate}
    \end{proposition}

    \begin{proof}
        \begin{enumerate}[label=\roman*)]
            \item \((T,V)\) absolutely irreducible and f.d. \(\implies T(A) = \operatorname{End}_K(V)\) is hence commutative. But \(\operatorname{End}_K(V) \cong M_{\dim_K(V)}(K) \implies [V:K] = 1\).
            
            \item a \(\implies\) b by i and the definition of splitting fields.
            
            b \(\implies\) because any one dimensional representation is absolutely irreducible.
        \end{enumerate} 
    \end{proof}

    These statements apply in particular to the case \(A = KG\) of a commutative group \(KG\).

    \begin{remark*}
        Infinite groups can have finite dimensional irreducible representations.
    \end{remark*}


    For example, consider \(G = \mathbb{Z}\) acting on \(\mathbb{C}^{(\mathbb{Z})} = \oplus_{n\in \mathbb{Z}} \mathbb{C} e_n\) . 

    \begin{definition}
        Let \(G\) be any irrep and \((T,V) \in \operatorname{Rep}_K(G)\).

        \begin{enumerate}[label=\roman*)]
            \item If \(\vert G \vert < \infty\), we call \(N_G = \sum_{g\in G=1} g \in KG\) the \underline{norm element} of \(KG\).
            
            The map \(v \mapsto T(N_G)(v) = \sum_{g\in G} T(g) \cdot v\) is called the \underline{norm map} on \(V\).
            
            \item The subspace \(H^o(G,V) \coloneqq V^G \coloneqq \{ v\in V : \forall g\in G, T(g) \cdot v = v \}\) is callled the fixed module of \(V\).
        \end{enumerate} 
    \end{definition}

    \begin{remark*}

        \begin{enumerate}[label=\arabic*)]
            \item \(V^G\) is itself a \(G\)-representation: every \(g\in G\) acts trivially (i.e. as identity) on \(V^G\). It is thus the direct sum of \([V^G : K]\) many copies of the trivial representation \(\mathbf{1}\).
            
            \(\mathbf{1}\) is the trivial \(1\)-dimensional vector space \(K\) with trivial group action.
            
            \item In the case of a finite group, \(\forall h\in G : h \cdot N_G = \sum_{g\in G} hg = \sum_{g\in G} g = N_G = N_G \cdot h\) in \(KG\).
            
            \item \(\forall v\in V: N_G \cdot v (= T(N_G) \cdot v) \in V^G\). \(h(N_G \cdot v) = (hN_G) \cdot v = N_G \cdot v\).
            \item \(\operatorname{Hom}_K(V,W)\) is naturally a \(G\)-rep for any \(V,W \in \operatorname{Rep}_K(G)\).
            
            \(\varphi : V \to W, (g \cdot \varphi)(v) = g \cdot \varphi (g ^{-1} \cdot v)\). We set \(V^{\ast} = \operatorname{Hom}_K(V,K)\). Then, \(\operatorname{Hom}_K(V,W)^G = \operatorname{Hom}_{KG}(V,W)\) [HW8].
            
            \item The assignment \(V \rightsquigarrow V^G = H^o(G,V)\) is a functor \(\operatorname{Rep}_K(G) \to K\text{-Mod}\) [= cat. of \(K\)-v.s.]
            
            i.e. there is a map \(\operatorname{Hom}_G(V,W) \to \operatorname{Hom}_K(V^G , W^G)\) given by \(\alpha \mapsto [v \mapsto \alpha(v)] = \alpha^G\) such that whenever \(V \xrightarrow{\alpha} W \xrightarrow{\beta} U\) in \(\operatorname{Rep}_K(G)\) we have \((\beta \circ \alpha)^G = \beta^G \circ \alpha^G\). \((\operatorname{id}_V)^G = \operatorname{id}_{V^G}\).  
        \end{enumerate} 
    \end{remark*}

    \begin{theorem}
        \begin{enumerate}[label=\roman*)]
            \item \(H^o(G,-)\) is left-exact. i.e. for any exact sequence:
            
            \[
                0 \to X \xrightarrow{\alpha} Y \xrightarrow{\beta} Z \to 0 \;. \tag*{$(\ast)$}
            \]

            in \(\operatorname{Rep}_K(G)\), the sequence:

            \[
                0 \to X^G \to Y^G \to Z^G\;.
            \]

            is exact.

            \item If \(\vert G \vert < \infty\) and \(\operatorname{char}(K) \nmid \vert G \vert\), then \(H^o(G,-)\) is an exact functor, i.e. for any exact sequence \((\ast)\), the sequence \(0 \to X^G \to Y^G \to Z^G \to 0\) is exact.
        \end{enumerate} 
    \end{theorem}

    \begin{proof}
        \begin{enumerate}[label=\roman*)]
            \item \(\alpha^G\) is clearly injective.
            
            \(\beta \circ \alpha = 0 \to \beta^G \circ \alpha^G = (\beta \circ \alpha)^G = 0^G = 0 \implies \operatorname{im}(\alpha^G) \subseteq \ker(\beta^G)\).

            Let \(y\in Y^G\) such that \(\beta(y) = 0\). Then \(y\in \ker(\beta) = \operatorname{im}(\alpha) \implies \exists x\in X : \alpha(x) = y\).

            \(\implies \forall g\in G, \alpha(gx) = g \cdot \alpha(x) = g\cdot y = y = \alpha(x)\). \(\alpha\) injective \(\implies \forall g\in G:  gx = x \implies x \in X^G\). We then have \(\alpha^G(x) = \alpha(x) = y\), thus \(\ker(\beta^G) \subseteq \operatorname{im}(\alpha^G)\).
            
            Therefore, \(\operatorname{im}(\alpha^G) = \ker(\beta^G)\) which proves the exactness at \(Y^G\).

            \item We need to prove that \(\beta^G\) is surjective. Let \(z\in Z^G\). \(\beta\) is surjective, so \(\exists y_0 \in Y\) such that \(\beta(y_0) = z\).
            
            Let \(y \in \frac{1}{\vert G \vert} \sum_{g\in G} g \cdot y_0\). Then, \(y\in Y^G\), and \(\beta(y) = \frac{1}{\vert G \vert} \sum_{g\in G} g \cdot \beta(y_0) = \frac{1}{\vert G \vert} \sum_{g\in G} g \cdot z = \frac{1}{\vert G \vert} (\vert G \vert \cdot z) = z\). 
        \end{enumerate} 
    \end{proof}

    \begin{remark*}
        \begin{enumerate}[label=\arabic*)]
            \item If \(\operatorname{char}(K) \nmid \vert G \vert < \infty\), then \(V = \ker(N_G : V\to V) \oplus V^G\) is a direct sum decomposition of \(G\)-representations.
            
            \begin{proof}
                Let \(\widetilde{N} = \frac{1}{\vert G \vert} N_G\). Then \(\operatorname{im}(\widetilde{N}) = V^G\) [since \(\widetilde{N} \cdot v = v\) for all \(v\in V^G\).] If \(v\in \ker(\widetilde{N}) \cap \operatorname{im}(\widetilde{N})\) then \(0 = \widetilde{N} v, v = \widetilde{N} w \implies 0 = \widetilde{N}^2 w = \widetilde{N} w = v\) where \(v= \widetilde{N}w\) for some \(w\in V\). This is because \(\widetilde{N}^2 = \frac{1}{\vert G \vert^2} \sum_{g_1 , g_2 \in G} g_1 g_2 = \frac{1}{\vert G \vert^2} \sum_{h \in G} \left( \sum_{(g_1 , g_2) \in G \times G , g_1 g_2 = h} 1 \right) h = \frac{1}{\vert G \vert^2} \sum_{h\in G} \vert G \vert h = \frac{1}{\vert G \vert} \sum_{h\in G} h = \widetilde{N}\).    
            \end{proof}

            \item The condition \(\operatorname{char}(K) \nmid \vert G \vert\) is \underline{necessary} for \(H^o(G,-)\) to be exact.
            
            For example, let \(G = \mathbb{Z} / 2 \mathbb{Z} = \{ \overline{0}, \overline{1} \} \) and \(K\) is any field of \(\operatorname{char} 2\). Let \(Y\) in the exact sequence by \(Ke_1 \oplus Ke_2\). Let \(X = Ke_1, Z = Y / X \cong K \overline{e_2}\) where \(T: G \to \operatorname{GL}(Y) = \operatorname{GL}_2(K)\) is given by \(\overline{0} \mapsto \begin{bmatrix}
                1 & 0 \\
                0 & 1 \\
            \end{bmatrix}\) and \(\overline{1} \mapsto \begin{bmatrix}
                1 & 1 \\
                0 & 1 \\
            \end{bmatrix}\). Note that in characteristic \(2\) it is indeed a representation.

            Note that, under this representation, \(e_2 \mapsto e_1 + e_2\). \(e_2\) is not fixed by the group action, but the image of \(e_2\) is fixed by the group action:

            \[
                \begin{bmatrix}
                    1 & 1 \\
                    0 & 1 \\
                \end{bmatrix} \overline{e_2} = \overline{e_1 + e_2} = \overline{e_2}.
            \]

            \[
                0 \to \underset{\cong \mathbf{1}}{X} \to Y  \to \underset{\cong \mathbf{1}}{Z} \to 0\;. 
            \]

            \[
                0 \to X^G = X \xrightarrow{\operatorname{id}_{X}} Y^G = X \to Z^G = Z = Y / X\;.
            \]

            Thus \(\beta^G\) must be the zero map, which is not surjective.

        \end{enumerate} 
    \end{remark*}

    \section*{Thursday, 3/26/2026}
    
    \begin{remark*}
        \(\operatorname{char} (K) \nmid \vert G \vert < \infty \implies V = V^G \oplus \ker (\widetilde{N}_G)\). Here \(\widetilde{N}_G = \frac{1}{\vert G \vert} N_G : V \to V\).
    \end{remark*}

    \begin{proposition}
        Let \(V, W \in \operatorname{Rep}_K(G)^{f.d.}\).

        \begin{enumerate}[label=\roman*)]
            \item \(\forall g\in G: \chi_{V^{\ast}}(g) = \chi_V(g ^{-1})\) [here \(V^{\ast} = \operatorname{Hom}_K(V,K)\)].
            \item Let \(\varphi \in \operatorname{Hom}_K(V,W)\). Recall that \((g \circ \varphi)(v) = g \cdot \varphi(g ^{-1} \cdot v)\). Thus, \(\chi_{\operatorname{Hom}_K(V,W)}(g) = \chi_V(g ^{-1}) \chi_W(g)\). 
        \end{enumerate} 
    \end{proposition}

    \begin{proof}
        \begin{enumerate}[label=\roman*)]
            \item HW8.
            \item \(V^{\ast} \otimes_K W \xrightarrow{\cong} \operatorname{Hom}_K (V,W)\) given by \(\lambda \otimes w \mapsto [v \mapsto \lambda(v)w]\) is an isomorphism of \(G\)-reps given \([V:K] < \infty\). Therefore, \(\chi_{\operatorname{Hom}_K(V,W)}(g) = \chi_{V^{\ast} \otimes_K W}(g) \overset{3.2}{=} \chi_{V^{\ast}}(g) \chi_W(g) = \chi_V(g ^{-1}) \chi_W(g)\).
        \end{enumerate} 
    \end{proof}

    \begin{proposition}
        Assume \(\operatorname{char}(K) \nmid n \coloneqq \vert G \vert < \infty\). Let \(V \in \operatorname{Rep}_K(G)^{f.d.}\).

        Then, \([V^G:K] \cdot 1_K = \frac{1}{n} \sum_{g\in G} \chi_V(g) = \frac{1}{n} \chi_V(N_G) = \chi_V(\widetilde{N}_G)\). 
    \end{proposition}

    \begin{proof}
        Remark after 3.9 \(\implies V = V^G \oplus \ker(\widetilde{N}_G)\). Let \(V_0 \coloneqq \ker(\widetilde{N}_G)\).

        Then, \(\widetilde{N}_G v = v\) for all \(v\in V^G\). Thus, \(\eval{\widetilde{N}_G}_{V^G} = \operatorname{id}_{V^G}, \eval{\widetilde{N}_G}_{V^0}=0_{V_G}\).
        
        Therefore, \([V^G:K]\cdot 1_K \equiv  \Tr(\operatorname{id}_{V_G}) = \Tr(\widetilde{N}_G : V \to V) = \frac{1}{n} \sum_{g} \Tr(T(g): V \to V) = \frac{1}{n} \sum_{g} \chi_V(g)\). 
    \end{proof}

    \begin{theorem}
        Suppose \(\operatorname{char}(K) \nmid n \coloneqq \vert G \vert < \infty\). Then every subrepresentation \(W\) of a representation \(V\) is a \underline{direct summand}.

        Equivalently, every submodule \(W\) of a \(KG\)-module \(V\) is a direct summand.

        Hence, every \(G\)-representation on a \(K\)-vector space is semisimple.

        Key: we don't need finite dimensional representation for this theorem.
    \end{theorem}

    \begin{proof}
        Let \(W \subset V\) be a \(G\)-subrepresentation. Consider the exact sequence of \(K\)-vector spaces:

        \[
            0 \to W \xrightarrow{\alpha=\text{inclusion}} V \xrightarrow[\text{canonical}]{\beta} Q \coloneqq V / W \to 0\;.
        \]

        Every short exact sequence of \(K\)-vector spaces splits. We can choose a section of \(\beta\): \(\sigma \in \operatorname{Hom}_K(Q, V)\) such that \(\beta \circ \sigma = \operatorname{id}_{Q}\).

        \underline{Claim}: \(\operatorname{Hom}_K(V,W) \xrightarrow{\alpha^{\ast}} \operatorname{Hom}_K(W,W) : \psi \mapsto \alpha^{\ast}(\psi) = \psi \circ \alpha = \eval{\psi}_{W}\) is surjective.

        \underline{Proof of Claim}: First note that \(\beta(v - \sigma(\beta(v))) = \beta(v) - \beta(\sigma(\beta(v))) = \beta(v) - \beta(v) = 0\).
        
        Therefore, \(v - \sigma(\beta(v)) \in \ker(\beta) = \operatorname{im}(\alpha) = W\).

        Given \(\varphi: W \to W\), set \(\psi : V \to W\) by \(\psi(v) = \phi(v - \sigma(\beta(v)))\).

        Then, \(\psi(\underset{\in W}{w}) = \varphi(w - \sigma(\beta(w))) = \varphi(w)\). So \(\varphi\) restricts to \(\psi\). Hence the restriction map is surjective.
        
        Now, Claim \(\implies 0 \to \operatorname{Hom}_K(Q,W) \xrightarrow[\lambda \mapsto \lambda \circ \beta]{\beta^{\ast}} \operatorname{Hom}_K(V,W) \xrightarrow{\alpha^{\ast}} \operatorname{Hom}_K(W,W) \to 0\) is an exact sequence.

        [\(\beta^{\ast}\) is injective, and the surjectivity of \(\alpha^{\ast}\) follows from the claim.]

        Theorem 3.9(ii) \(\implies 0 \to \operatorname{Hom}_K(Q,W)^G \to \operatorname{Hom}_K(V,W)^G \to \operatorname{Hom}_K(W,W)^G \to 0\) is exact.

        HW8 implies that \(\operatorname{Hom}_K(Q,W)^G = \operatorname{Hom}_{KG}(Q,W)\) etc. So we have the exact sequence:

        \[
            0 \to \operatorname{Hom}_{KG}(Q,W) \to \operatorname{Hom}_{KG}(V,W) \to \operatorname{Hom}_{KG}(W,W) \to 0
        \]

        By surjectivity, there must be a \(KG\)-module homomorphism \(\psi : V \to W\) such that \(\eval{\varphi}_{W} = \operatorname{id}_{W}\).

        \underline{Claim}: \(V = W \oplus \ker(\psi)\).

        \underline{Proof of Claim}: \(v = \psi(v) + (v - \psi(v))\). Note that, \(\psi(v-\psi(v)) = \psi(v) - \psi(\psi(v))\). \(\psi(v) \in W\) and \(\eval{\psi}_{W} = \operatorname{id}_{W} \implies \psi(\psi(v)) = \psi(v) \implies \psi(v-\psi(v)) = 0\). Therefore \(v - \psi(v) \in \ker(\psi)\).

        Now, suppose \(w\in W\cap \ker(\psi)\). \(w\in W \implies \psi(w) = w \implies w = 0\). Thus \(W \cap \ker(\psi) = 0\).

        Thus indeed \(V = W \oplus \ker(\psi)\).

    \end{proof}

    \begin{theorem}
        [Maschke's Theorem, 1899]

        Suppose \(G\) is a finite group of order \(n\). TFAE:

        \begin{enumerate}[label=\roman*)]
            \item \(KG\) is semisimple.
            \item \(\operatorname{char}(K) \nmid n\). 
        \end{enumerate} 
    \end{theorem}

    \begin{proof}
        ii \(\implies\) i: \((KG)_\ell\) is a semisimple \(KG\)-module by Theorem 3.12.

        i \(\implies\) ii: Suppose \(KG\) is semisimple but \(p \coloneqq \operatorname{char}(K), p\mid n\).
        
        Consider \(x \coloneqq N_G = \sum_{g\in G} g \in KG\).

        \(\forall h\in G: h \cdot N_G = h \cdot \sum_{g\in G} g = \sum_{g\in G} h \cdot g = \sum_{g\in G} g = N_G\).

        So, \(x\in Z(KG)\). \(Z(KG)\) must be a finite product of fields [Corollary 2.7].

        On the other hand, \(x^2 = \sum_{(g_1 , g_2) \in G \times G} g_1 g_2 = \sum_{h\in G} \left( \sum_{(g_1, g_2) \in G \times G , g_1 g_2 = h} 1 \right) h = \sum_{h\in G} n h = 0\). 

        In a product of fields, \(x^2 = 0\) has solution \(x = 0\). But \(\sum_{g\in G} g \neq 0\), which is a contradiction.
    \end{proof}

    \section*{Tuesday, 3/31/2026}
    
    \begin{theorem}
        Suppose \(\operatorname{char}(K) \nmid \vert G \vert = n < \infty\). Let \(A = KG = A_1 \times \cdots \times A_k.\). \(\forall 1 \leq i \leq k\) let \((T_i , V_i)\) be a simple \(A_i\)-module.

        \begin{enumerate}[label=\roman*)]
            \item Each simple \(KG\)-module is isomorphic to exactly one of these \(V_i\). 
            \item If \(D_i = \operatorname{End}_{A_i}(V_i)\) and \((A_i)_\ell \cong V_i^{\oplus n_i}\) then,
            
            \(A_i = \operatorname{End}_{A_i}((A_i)_\ell)^o \cong \operatorname{End}_{A_i}(V^{\oplus n_i})^o = M_{n_i}(\operatorname{End}_{A_i}(V_i))^o = M_{n_i}(D_i)^o = M_{n_i}(D_i^o)\).

            \item If \(e_i = 1_{A_i} \implies 1_A = e_1 + \cdots + e_k , e_i e_j = \delta_{ij} e_i\) [central idempotents], \(T_i(e_j) = \delta_{ij} \operatorname{id}_{V_i}\).
            \item Set \(d_i = [V_i : K] , K_i = Z(A_i) = Z(D_i), s_i = s(D_i)\) [so \(s_i^2 = [D_i : K_i]\)], then \(d_i = n_i [D_i : K] = n_i s_i^2 [K_i : K]\) and \(\boxed{(1)\; n = \sum_{i=1}^k n_i^2 s_i^2 [K_i : K]}\).
            \item If \(\chi_i \coloneqq \chi_{V_i}\) then \(\chi_i(e_j) = \delta_{ij} \cdot d_i \cdot 1_K\), and \(\boxed{\rho_G \coloneqq \rho_{(KG)_\ell} \overset{(2)}{=} \sum_{i=1}^k n_i \chi_i}\).
            \item Assume \(\operatorname{char}(K) = 0\). Then, \(V,W \in \operatorname{Rep}_{K}(G)^{\text{f.d.}}\) are isomorphic \(\iff \chi_V = \chi_W\). 
        \end{enumerate} 
    \end{theorem}    

    \begin{proof}
        \begin{enumerate}[label=\roman*)]
            \item Follows from ch. 2. [HW2/6]

            \item The isomorphism \(A_i \xrightarrow{\cong} M_{n_i}(D^o)\) follows from the proof of Wedderburn's theorem.
             

            \item If \(e_i = 1_{A_i} \implies 1_A = e_1 + \cdots + e_k , e_i e_j = \delta_{ij} e_i\) [central idempotents]. \(T_i(e_j) = \delta_{ij} \operatorname{id}_{V_i}\).
            
            \item ii \(\implies [A_i : K] = n_i^2 [D_i : K] = n_i^2 [D_i : K_i][K_i : K] = n_i^2 s_i^2 [K_i : K]\). \([A_i : K] = [V_i^{\oplus n_i} : K] = n_i [V_i : K]\). Cancel \(n_i \implies [V_i : K] = n_i s_i^2 [K_i : K]\) and \(n = [A:K] = \sum_{1 \leq i \leq k} [A_i : K] = \sum_{1 \leq i \leq k} n_i^2 s_i^2 [K_i : K]\). 
            
            \item \(\chi_i(e_j) = \Tr(T_i(e_j): V_i \to V_i) = \begin{dcases}
                0, &\text{ if } j\neq i ;\\
                [V_i : K] \cdot 1_K, &\text{ if } j=i.
            \end{dcases}\) 

            \((KG)_\ell = A_\ell = (A_1)_\ell \times \cdots \times (A_k)_\ell \cong V_1^{\oplus n_1} \oplus \cdots \oplus V_k^{\oplus n_k} \implies \rho_G = \chi_{(KG)_\ell} = \sum_{i=1}^k n_i \chi_i\).

            \item \(KG\) semisimple, so \(V \cong V_1^{\oplus n_1} \oplus \cdots \oplus V_k^{\oplus n_k}, W = V_1^{\oplus \ell_1} \oplus \cdots \oplus V_k^{\oplus \ell_k}\).
            
            \(V \cong W \iff \forall 1 \leq i \leq k : m_i = \ell_i \iff  \sum_{i=1}^k m_i \chi_i = \sum_{i=1}^k \ell_i \chi_i\). Apply both sides to \(e_j: m_j \chi_j(e_j) = \ell_j \chi_j(e_j)\). Since \(\chi_j(e_j) = d_j \cdot 1_K\), \(\operatorname{char}(K) = 0 \implies m_j = \ell_j\). 

        \end{enumerate} 
    \end{proof}

    \begin{theorem}
        Suppose \(\operatorname{char}(K) \nmid \vert G \vert = n < \infty\) and let \(KG = A = A_1 \times \cdots \times A_k\) and \((T_i , V_i), 1 \leq i \leq k\), as in 3.14. Then TFAE:
        
        \begin{enumerate}[label=\roman*)]
            \item \(K\) is a splitting field of \(G\) [every f.d. irreducible representation of \(G\) over \(K\) is absolutely irreducible.]
            \item \(\forall 1 \leq i \leq k : (T_i , V_i)\) is absolutely irreducible.
            \item \(\forall 1 \leq i \leq k : D_i = \operatorname{End}_A(V_i) = \operatorname{End}_{A_i}(V_i) = K\).
            \item \(\forall 1 \leq i \leq k : A_i \cong M_{\ell_i}(K)\) for some \(\ell_i\). 
            \item \(n = \vert G \vert = \sum_{1 \leq i \leq k} d_i^2 , d_i = [V_i : K]\).
        \end{enumerate} 
    \end{theorem}

    \begin{proof}

        \ \

        \begin{enumerate}
            \item[i \(\iff\) ii:] Every irreducible representation of \(G\) over \(K\) is one of \(V_1 , \cdots , V_k\) by 3.14(i).
            \item[ii \(\iff\) iii:] 3.7 : \(V_i\) is absolutely irreducible \(\iff \operatorname{End}_A(V_i) = K\).
            \item[iii \(\iff\) iv:] \(A_i = M_{n_i}(D_i^o) = M_{\ell_i}(K) \overset{\text{Wedderburn}}{\iff} K = D_i^o [\iff K = D_i]\).
            \item[v \(\iff\) iii] \(n \overset{3.14(vi)}{=} \sum_{i=1}^k n_i^2 s_i^2 [K_i : K] \overset{d_i^2 = n_i^2 s_i^4 [K_i : K]^2}{\leq} \sum_{i=1}^n d_i^2 \overset{(v)}{=} n\). Equality holds \(\iff \forall 1 \leq i \leq k: n_i^2 s_i^4 [K_i : K]^2 = n_i^2 s_i^2 [K_i : K]\). Cancelling, \(\iff s_i = 1, [K_i : K] = 1\).
            
            \(s_i = 1\) means \(D_i\) is its center, and \([K_i : K] = 1\) means \(K_i = K\), so this is equivalent to \(D_i = K\).
        \end{enumerate}
    \end{proof}

    Examples:

    \begin{enumerate}[label=\arabic*)]
        \item \(\operatorname{char}(K) \nmid 6 \overset{\text{HW 8}}{\implies} K[\mathscr{S}_3] \cong \underset{\textbf{1}}{K} \times \underset{\text{sgn}}{K} \times M_2(K)\), where \(M_2(K)\) is the irreducible 2\(\)-dimensional representation.
        
        \(6 = 1^2 + 1^2 + 2^2\).
        
        \item \(G = \mathbb{Z} / n\mathbb{Z}, K = \mathbb{C} \text{ or } K = \mathbb{Q}(\zeta_n), \zeta_n =\) primitive \(n\)'th root of unity, or \(K = \mathbb{F}_q\) with \(n\mid q-1\).
        
        Then \(KG \cong K[T] / (T^n - 1) \cong \prod_{i=0}^{n} K[T] / (T - \zeta_n^i)\) and \(K[T] / (T - \zeta_n^i) \cong K\) so \(KG \cong \underbrace{K \times \cdots \times K}_{n \text{ times}}\).
        
        \(n = 1^2 + \cdots + 1^2\).

        Representations correspond to \((T \mapsto 1), (T \mapsto \zeta_n) , (T \mapsto \zeta_n^2) , \cdots , (T \mapsto \zeta_n^{n-1})\).

        \item \(G = D_n\), dihedral group of order \(2n\), which is \(\mathbb{Z} / n\mathbb{Z}  \rtimes \mathscr{S}_2\) with \(\sigma \in \mathscr{S}_2\), \(\sigma = [\overline{a} \mapsto -\overline{a}]\).
        
        Irreducible representations of \(D_n\) over \(\mathbb{C}\) are \(1\) or \(2\)-dimensionals. Proof (HW10).
        
        Check \(2n = \underbrace{1^2 + \cdots + 1^2}_{1\text{-dim  reps}} + \underbrace{2^2 + \cdot + 2^2}_{2\text{-dim reps}}\), so we can tell when we have exhausted finding all such representations. 

        All \(s_i = 1, [K_i : K] = 1\).
    \end{enumerate} 

    \begin{theorem}
        Suppose \(\operatorname{char}(K) \nmid n = \vert G \vert < \infty\) and \(G\) is abelian. TFAE:

        \begin{enumerate}[label=\roman*)]
            \item \(K\) is a splitting field of \(G\).
            \item Every irreducible representations of \(G\) has degree \(1\).
            \item \(G\) has exactly \(n\) inequivalent irreducible representations over \(K\). 
        \end{enumerate} 
    \end{theorem}

    \begin{proof}
        Combine 3.15 and 3.8
    \end{proof}

    Example: \(G = \mathbb{Z} / n_1 \mathbb{Z} \times \cdots \times \mathbb{Z} / n_s \mathbb{Z}\).

    Any \(1\)-dimensional representation of \(G\) over \(K\) is of the form:

    \[
        \chi(\overline{a_1}, \cdots , \overline{a_s}) = \zeta_{n_1}^{a_1} \cdots \zeta_{n_s}^{a_s}
    \]

    where \((\zeta_{n_i})^{n_i} = 1\).

    \(K\) is a splitting field of \(G \iff K\) contains primitive roots of unity of orders \(n_1 , \cdots , n_s\). Which is equivalent to containing one primitive root of unity of order \(\operatorname{lcm}(n_1 , \cdots , n_s)\).

    \(\iff \#\) of absolutely irreducible representations of \(G\) over \(K = n_1 \times \cdots \cdots \times n_s = n\).

    There are \(n_j\) choices, since we have \(n_j \leftrightarrow [\sigma \mapsto \zeta_{n_j}^{f_i}, 0 \leq f_i \leq n_j]\). 

    \begin{theorem}
        Suppose \(\operatorname{char}(K) \nmid \vert G \vert < \infty\). Denote by \(k = k(G,K)\) the number of inequivalent irreducible representations of \(G\) over \(K\). Let \(h\) be the number of conjugacy classes in \(G\). \(h\) is also often called the class number of \(G\).

        \begin{enumerate}[label=\roman*)]
            \item \(k \leq h\).
            \item TFAE: \begin{enumerate}[label=\alph*)]
                \item \(k = h\).
                \item \([Z(KG):K] = k\).
                \item \([Z_K(G):K] =  k\), \(Z_K(G)\) class functions.
                \item \(Z(KG)\) is a product of copies of \(k\).   
            \end{enumerate}
            
            \item \(K\) is a splitting field of \(G\), then \(k = h\).
            \item If \(\operatorname{char}(K) > 0\) and \(k = h\), then \(K\) is a splitting field of \(G\).
        \end{enumerate} 
    \end{theorem}

    \begin{proof}
        Write \(KG \overset{(1)}{=} A_1 \times \cdots \times A_k\) as in 3.14, \(A_i \cong M_{n_i}(D_i^o), Z(D_i) \eqqcolon K_i\).

        Theorem 3.4 \(\implies [Z(KG):K] \overset{(2)}{=} h\).

        \begin{enumerate}[label=\roman*)]
            \item (1) \(\implies Z(KG) = Z(A_1) \times \cdots \times Z(A_k) = K_1 \times \cdots \times K_k \implies \sum_{i=1}^k [K_i : K] \overset{(2)}{=} h \implies k \leq h\).
            \item (a): \(h = k \implies [Z(KG):K] = k\), hence b and c [since \([Z_K(G) : K] = h\)].
            
            \(h=k \iff \forall 1 \leq i \leq k : K_i = K \overset{(1)}{\iff} (KG)\) is a product of copies of \(K\) itself.

            \item \(\implies\): If \(K\) is a splitting field of \(G\), by 3.15, \(KG = M_{n_1}(K) \times \cdots \times M_{n_k}(K)\). Taking center, \(Z(KG) = K \times \cdots \times K\). Thus the conditions in ii are satisfied.
            
            \item Later.
            
        \end{enumerate} 
    \end{proof}

    \section*{Thursday, 4/2/2026}
    
    \subsection*{Example of group \(G\) and a field \(K\) with \(h\) irreducible inequivalent representations over \(K\), but \(K\) is \underline{not} a splitting field.}

    Let \(\mathbb{H} = \mathbb{R} \cdot 1 \oplus \mathbb{R} i \oplus \mathbb{R} j \oplus \mathbb{R} k\) be the quarternion algebra and let \(Q = \{ \pm 1, \pm i, \pm j, \pm k \}\).

    The `easiest' represnetations to think about for a group \(G\). first is the one-dimensional representation, which are homomorphisms \(G \to K^\times\). But \(K^\times\) is abelian, so such a homomorphism must factor through the abelianization of the group. So \(G^{\text{ab}} = [G,G]\) is important to consider.

    Here \(G=Q\), \([Q,Q] \ni ij i ^{-1} j ^{-1} = k (ji) ^{-1} = k(-k)^{-1}=-1\). Thus all commutators \(\in \{ 1, -1 \}\).
    
    \(Q / \{ \pm 1 \} \cong \mathbb{Z} / 2 \times \mathbb{Z} / 2\). So there are four one-dimensional represnetations.

    Note: \(KG\) semisimple \(\iff \operatorname{char}(K) \neq 2\) which we assumed.

    Let \(\mathbf{1}\) be the the trivial representation. The four representations are:
    
    \(V_1 = \mathbf{1} \otimes \mathbf{1}, V_2 =  \mathbf{1} \otimes \text{sgn}, V_3 = \text{sgn} \otimes \mathbf{1}, V_4 = \text{sgn} \otimes \text{sgn}\).

    \(\text{sgn}(1 \text{ mod } 2) = -1\).
    
    Identify \(\mathbb{C} = \mathbb{R} \cdot 1 \oplus \mathbb{R} i\) and \(\mathbb{H} = \mathbb{C} \oplus \mathbb{C} \cdot j\), which is possible since \(\mathbb{C} \cdot j = \mathbb{R} \cdot j \oplus \mathbb{R} \cdot (ij) = \mathbb{R} \cdot j \oplus \mathbb{R} \cdot k\). So \(\mathbb{H}\) is a \(2\)-dimensional \(\mathbb{C}\)-v.s.
    
    \(\mathbb{R} G = \mathbb{R} \times \mathbb{R} \times \mathbb{R} \times \mathbb{R} \times \mathbb{R} \times \mathbb{H}\), where each \(\mathbb{R}\) corresponds to each of of the four \(1\)-dimensional representations.

    Naming in order, \(\mathbb{R} G = A_1 \times A_2 \times A_3 \times A_4 \times A_5\) as in 3.14. Then \(\mathbb{H} = \operatorname{End}_G(V_5)\)

    The decomposition of \(\mathbb{R} G\) is not split, since \(\mathbb{H}\) is a division algebra not \(\mathbb{R}\). Hence \(\mathbb{R}\) is not a splitting field of \(G\).

    But \(Z(\mathbb{R} G) = \mathbb{R} \times \mathbb{R} \times \mathbb{R} \times \mathbb{R} \times \mathbb{R}\).

    \(G\) has \(5\) conjugacy classes: \(1, -1, i,j,k\).

    \subsection*{The Semisimplicity Hypothesis (SSH)}

    We say a group has SSH if \(\operatorname{char}(K) \nmid \vert G \vert \eqqcolon n < \infty\). \(n\) will be always used as the order of the group.

    \begin{theorem}
        Assume SSH. Let \(A = KG = A_1 \times \cdots \times A_k\) as in 3.14 with \(e_i = 1_{A_i}\).
        
        Then, \(e_i = \frac{(n_i \cdot 1_K)}{(n\cdot 1_K)} \sum_{g\in G} \chi_i(g ^{-1}) \cdot g \in KG\).
        
        Here, \(n_i = \frac{[V_i : K]}{[D_i : K]}\), as in 3.14. As \(e_i \neq 0\), we find that \(n_i \cdot 1_K \neq 0_K\).

        If \(K\) is a splitting field of \(G\), then \(D_i = K\) and thus \(n_i = [V_i : K]\) for \(1 \leq i \leq k\).
    \end{theorem}

    \begin{proof}
        We can write \(e_i = \sum_{x\in G} a_x x\), \(a_x\in K\). 


        \(\forall g\in G\), let the character of the regular representation be \(\rho_G\). Then, \(\rho_G(g ^{-1} e_i) = \sum_{x\in G} a_x \rho_G(g ^{-1} x)\).

        Note that \(\rho_G(g ^{-1} x) = n\) if \(x = g\), \(0\) otherwise. So, \(\rho_G(g ^{-1} e_i) = a_g \cdot n\;(1)\).

        3.14(v) \(\implies \rho_G = \sum_{i=1}^k n_i \chi_i\).
        
        Then \(\rho_G(g^{-1} e_i) = \sum_{j=1}^k n_j \chi_j(g^{-1} e_i) = n_i \chi_i (g^{-1})\;(2)\).
        
        Since \(A \twoheadrightarrow A_j\) and then \(A_j\) acts on \(V_j\).

        (1) and (2) together imply \(n a_g = n_i \chi_i(g^{-1})\). Dividing by \(n\), we see: \(a_g = \frac{n_i}{n}\chi (g ^{-1})\).


    \end{proof}

    Now we show the proof of 3.17(iv).

    \begin{proof}
        By 3.18, \(e_i = \frac{n_i}{n} \sum_{g\in G} \chi_i(g ^{-1}) g\).

        Lemma 3.6 implies that the coefficients are sums of roots of unity, which implies that \(\chi_i(g ^{-1})\) are algebraic over \(\mathbb{F}_p\) where \(p = \operatorname{char}(K)\).
        
        Let \(F / \mathbb{F}_p\) be a finite field extension which contains all the values \(\chi_i(g)\) for all \(1 \leq i \leq k\). \(F \subseteq K \implies e_i \in FG \subset KG\).

        Then \(FG = FG_1 \oplus FG_2 \oplus \cdots \oplus FG_n\), where \(FG_i = e_i FGe_i \subset FG \subset KG\) is a non-zero \(F\)-subalgebra of \(FG\).

        \(FG_i \otimes_F K \subset e_i KG e_i = A_i \implies [FG_i : F] \leq [A_i : K]\). Note that \([FG_i : F] \equiv  [FG_i \otimes_F K : K]\). Take \(\sum_{i}\) and compare, the inequality is an equality. Thus \(FG_i \otimes_F K = A_i\).
        
        Therefore, \(FG_i\) is simple. Wedderburn 2.11 \(\implies FG_i \cong M_{\ell_i}(\text{division algebra})\). \(F\) is a finite field, so the division algebra is some field extension of \(F\).

        Thus \(FG_i \cong M_{\ell_i}(F_i)\).

        \(Z(FG) = Z(FG_1) \times \cdots \times (FG_h) = F_1 \times \cdots \times F_h\).

        But \(k(G,F) = h\). 3.17(ii) implies that \([Z(FG):F] = h\). Then each field extension must be the trivial extension. i.e. \(F_i = F\) for all \(i\). This implies that indeed, the group algebra is split.

        By 3.15(iii), \(F\) is a splitting field. By extending scalars, \(K\) is a splitting field.
    \end{proof}

    \begin{definition}
        Assume SSH. For \(\lambda , \mu \in Z_K(G)\), any two class functions, we define a pairing as follows:

        \[
            \langle \lambda , \mu \rangle \coloneqq \frac{1}{n} \sum_{x\in G} \lambda(x) \mu(x ^{-1}) \in K
        \]

        Note that \(\langle \lambda , \mu \rangle = \langle \mu , \lambda \rangle \).
    \end{definition}
    
    \begin{theorem}
        [Orthogonality Relations]

        Assume SSH. Let \(\chi_1 , \cdots , \chi_k\) be the characters of the pairwise inequivalent irreducible reps \(V_1 , \cdots , V_k\) over \(K\).

        \begin{enumerate}[label=\roman*)]
            \item \(\forall 1 \leq i, j \leq k\),
            
            \begin{enumerate}[label=\alph*)]
                \item \(\frac{1}{n} \sum_{x\in G} \chi_i(xg) \chi_j(x ^{-1}) = \begin{dcases}
                    \frac{1}{n_i}\chi_i(g), &\text{ if } j=i ;\\
                    0, &\text{ if } j\neq i.
                \end{dcases}\)  
                \item \(\langle \chi_i , \chi_j \rangle \coloneqq \frac{1}{n} \sum_{x\in G} \chi_i(x) \chi_j(x ^{-1}) = \delta_{ij} [D_i : K] \cdot 1_K\).
            \end{enumerate} 

            \item If \(K\) is a splitting field of \(G\), then \(\forall 1 \leq i , j \leq k : \langle \chi_i, \chi_j \rangle = \delta_{ij} \cdot 1_K\).
        \end{enumerate} 
    \end{theorem}

    \begin{proof}
        \(e_i \overset{3.18}{=} \frac{n_i}{n} \sum_{x\in G} \chi_i(x) x ^{-1} = \frac{n_i}{n} \sum_{x\in G} \chi_i(xg)(xg)^{-1} = g ^{-1} \frac{n_i}{n} \sum_{x\in G} \chi_i(xg) x ^{-1}\).

        Thus, \(g e_i = \frac{n_i}{n} \sum_{x\in G} \chi_i(xg) x ^{-1}\).

        Apply \(\chi_j\). Then,

        \(\frac{n_i}{n} \sum_{x\in G} \chi_i(xg) \chi_j(x ^{-1}) =\chi _j(ge_i)\).
        
        \(ge_i \in A_i \implies \chi_j(ge_i) = \begin{dcases}
            0, &\text{ if } j\neq i ;\\
            \chi_i(g), &\text{ if } j=i.
        \end{dcases}\)

        Hence (a).

        For (b), evaluate (a) at \(g=1_G\).

        We get: \(\langle \chi_i , \chi_i \rangle = \frac{1}{n_i} \chi_i(1_G) = \frac{[V_i : K]}{n_i} = \frac{[V_i : D_i] \cdot [D_i : K]}{n_i} = [D_i : K]\) 

        ii follows from 3.15(iii).
    \end{proof}

    \begin{theorem}
        Assume SSH. Let \(m\) be the exponent of \(G\).

        If \(K\) contains a primitive \(m\)'th root of unity, then \(k(G,K) = h\), and hence the equivalent statements in 3.17(ii) hold.
    \end{theorem}
    
    \begin{remark*}
        In fact, if \(K\) contains a primitive \(m\)'th root of unity, then \(K\) is a splitting field of \(G\). If \(\operatorname{char}(K) > 0\), we have seen this in 3.17(iv). If \(\operatorname{char}(K) = 0\), then it will be shown later.
    \end{remark*}

    \begin{proof}
        Similar to 3.17(iv). HW10.
    \end{proof}
\end{document}